### 2018-03-29 inst/manuscript/manuscript_supplemental.Rmd
adding manuscript supplemental to repo

---
title: "_Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions."
author: "Nate Olson"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
---

## Background
16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance analysis. 
A limitation to 16S rRNA amplicon sequencing is a lack of taxonomic resolution, where organisms are only identifiable to the genus or family level. 
We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus. 
Taxonomic resolution can vary by clade and amplicon regions, though the extent to which it varies is not well characterized.

Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions 
Using the Greengenes 13.5, `greengenes13.5MgDb` we will first subset the database to obtain _Paenibacillus_ taxonomic and sequence data. 
Next we will explore the _Paenibacillus_ species composition, then characterize the within and between species sequence similarity for the 16S rRNA gene V12 and V4 regions.

## Required Packages  
Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix. 
The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures.

```{r message = FALSE, echo = TRUE}
library(tidyverse); packageVersion("tidyverse")
library(ggpubr); packageVersion("ggpubr")
library(DECIPHER); packageVersion("DECIPHER")
library(metagenomeFeatures); packageVersion("metagenomeFeatures")
library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
```

## _Paenibacillus_ Sequence and Taxonomy Data
We first subset the Greeengenes 13.5 database using the `mgDb_select` function. 
Then summarize the taxonomy data using the `tidyverse` package. 
<!-- Loading metagenomeFeatues and greengenes db -->

```{r}
paeni_16S <- mgDb_select(gg13.5MgDb, 
                          type = c("taxa","seq"), 
                          keys = "Paenibacillus", 
                          keytype = "Genus")
```

```{r}
## Per genus count data
taxa_df <- paeni_16S$taxa %>% 
    ## cleaning up species names
    mutate(Species = if_else(Species == "s__", "Unassigned", Species),
           Species = str_replace(Species, "s__","")) %>% 
    group_by(Species) %>% 
    summarise(Count = n()) %>% 
    ungroup() %>% 
    mutate(Species = fct_reorder(Species, Count))

## Count info for text 
total_otus <- sum(taxa_df$Count)
no_species_assignment <- taxa_df$Count[taxa_df$Species == "Unassigned"]
```

For the Greengenes 13.5 database there are a total of `r total_otus` sequences classified as `r nlevels(taxa_df$Species)` species in the Genus _Paenibacillus_. 
The number of sequences assigned to specific _Paenibacillus_ species, range from 199 for _Paenibacillus amylolyticus_ to 2 for _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)). 
Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment`. 
```{r speciesCount, fig.cap = "Number of sequences assigned to species in the genus _Paenibacillus_."} 
taxa_df %>% 
ggplot() + 
    geom_bar(aes(x = Species, y = Count), stat = "identity") + 
    geom_text(aes(x = Species, y = Count, label = Count),nudge_y = 75) +
    labs(y = "Number of OTUs") + 
    coord_flip() + 
    theme_bw() + 
    theme(axis.text.y = element_text(face = "italic"))
```

********************************************************************************

## Taxonomic resolution 
Next we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ species by comparing within and between species amplicon pairwise distance for the V12 and V4 regions. 
Inorder to differentiate between species the pairwise distances for within species amplicon seqences must be less than the between species distances. 
Additionally the difference in amplicon sequence pairwise distances between and within species must be greater than the sequencing error rate to detect the difference. 
For our taxonomic resolution analysis we use _in-silico_ PCR to extract the V12 and V4 regions of the 16S rRNA sequences. 
We will then generate a pairwise distance matrix for the two regions and compare the within and between species pairwise distances.

For our _in-silico_ PCR we will use the following PCR primers:

|Region |Forward                    |Reverse                        |  
|:------|:--------------------------|:------------------------------|  
|V12    |27F - AGAGTTTGATCATGGCTCAG |336R - CACTGCTGCSYCCCGTAGGAGTCT|  
|V4     |515F - GTGCCAGCMGCCGCGGTAA |806R - GGACTACHVGGGTWTCTAAT    |

### V12 
Extracting the V12 region from the database sequences, only sequences with the expected amplicon length of 318 basepairs are included in the analysis. 
```{r}
forward_primer <- "AGAGTTTGATCATGGCTCAG" 
## reverse complementing reverse primer
reverse_primer <- as.character(reverseComplement(DNAString("CACTGCTGCSYCCCGTAGGAGTCT")))

paeni_V12 <- TrimDNA(paeni_16S$seq, 
                               leftPatterns = forward_primer, 
                               rightPatterns = reverse_primer,
                               type = "both")

seqs_to_use <- (width(paeni_V12[[2]]) == 318)
v12_filtered <- paeni_V12[[2]][seqs_to_use]
```

Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.
```{r}

v12_align <- AlignSeqs(v12_filtered, verbose = FALSE)
```

The resulting alingment can be viewed using the `BrowseSeqs` function in the `DECIPHER` package.
```{r eval = FALSE}
BrowseSeqs(v12_align)
```

Generating pairwise distance matrix for taxonomic resolution analysis and converting distance matrix to data frame for analysis.

```{r}
v12_dist <- DistanceMatrix(v12_align, 
                           correction = "none", 
                           verbose = FALSE, 
                           includeTerminalGaps = FALSE)

v12_dist_df <- v12_dist %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "Keys") %>% 
    gather("Keys2","distance", -Keys) %>% 
    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>% 
    filter(Keys < Keys2) %>% 
    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))

tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species") 
v12_dist_anno_df <- v12_dist_df %>% 
    left_join(tax_df) %>% 
    left_join(tax_df,by = c("Keys2" = "Keys")) %>% 
    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>% 
        mutate(group_comp = if_else(Keys_Species == Keys2_Species, "within","between")) %>% 
    filter(Keys_Species != "s__", Keys2_Species != "s__")
```

### V4 
For the V4 region we will use the same approach, extract amplicon region, filter extracted sequences based on amplicon length, generate pairwise distance matrix using a multiple sequence alignment, and then evaluate pairwise distances.

```{r}
## Extract amplicon region
paeni_V4 <- TrimDNA(paeni_16S$seq, 
                               leftPatterns = "GTGCCAGCMGCCGCGGTAA", 
                               rightPattern = "ATTAGAWACCCBDGTAGTCC",
                               type = "both")

## Filter based on amplicon length
v4_filtered <- paeni_V4[[2]][width(paeni_V4[[2]]) == 252]

### Calculate distance matrix from multiple sequence alignment
v4_align <- AlignSeqs(v4_filtered, verbose = FALSE)
v4_dist <- DistanceMatrix(v4_align, 
                          correction = "none", 
                          verbose = FALSE, 
                          includeTerminalGaps = FALSE)

## Creating a data frame for exploratory analysis
v4_dist_df <- v4_dist %>% 
    as.data.frame() %>% 
    rownames_to_column(var = "Keys") %>% 
    gather("Keys2","distance", -Keys) %>% 
    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>% 
    filter(Keys < Keys2) %>% 
    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))

tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species") 
v4_dist_anno_df <- v4_dist_df %>% 
    left_join(tax_df) %>% 
    left_join(tax_df,by = c("Keys2" = "Keys")) %>% 
    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>% 
    mutate(group_comp = if_else(Keys_Species == Keys2_Species, "within","between")) %>% 
    filter(Keys_Species != "s__", Keys2_Species != "s__")

## Excluding outlier sequence "329842" - mean pairwise distance to all other
## sequences is 0.2
 v4_dist_anno_filt <- filter(v4_dist_anno_df,  Keys != "329842", Keys2 != "329842")
```

********************************************************************************

### Amplicon Sequence Lengths
The trimmed sequence length ranges between 0 and 1500, due to sequences without primer matches (Fig. \@ref(fig:trimInfo)). 
These are likey partial sequences or sequences with mutation in the primer binding regions. 
The R package `PrimerMatch` (__REF__) can be used evaluate mismatches in the primer binding regions. 
Here we will limit our taxonomic resolution analysis to trimmed sequences that are consistent with the expected amplicon length of 318 bp and 252 bp for the V12 and V4 regions repectively.

```{r trimInfo, fig.cap = "Primer trimmed sequence length and start and end positions relative to full length sequences for the V12 and V4 regions.", fig.height = 4}
list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>% 
    map_dfr(as.data.frame, .id = "amplicon") %>% 
    gather("key","value", -names, -amplicon) %>% 
    mutate(key = if_else(key == "width", "length", key),
           key = factor(key, levels = c("length","start","end"))) %>% 
    mutate(position_type = if_else(key == "length","position", "length")) %>% 
    ggplot() + 
    geom_histogram(aes(x = value, fill = key), bins = 100) + 
    facet_grid(position_type~amplicon, scales = "free") +
    theme_bw()
```

********************************************************************************

#### Genus Level Comparison

Pairwise distance is significantly different for within and between species comparisons indicating that the V12 and V4 regions can potentially be used to classify members of the _Paenibacillus_ genus to the species level (Fig. \@ref(fig:pairDist)).
Overall the V12 region had greater pairwise distances than V4 for both within and between species.
It is important to also consider that the majority of sequences in the database were only classified to the genus level. 
Species level information for these sequences might yield results that are inconsistent with our analysis.

```{r pairDist, fig.cap = "Distribution of within and between species pairwise distances for the V4 16S rRNA region. Sequences not classified to the species level were excluded from the analysis.", fig.height = 3}
list("V4" = v4_dist_anno_filt, "V12" = v12_dist_anno_df) %>% 
    bind_rows(.id = "amplicon")  %>% 
    ggplot(aes(x = group_comp, y = distance, fill = group_comp)) + 
    geom_boxplot(aes(x = group_comp, y = distance, fill = group_comp)) + 
      stat_compare_means(label.y = 0.2) +
    facet_wrap(~amplicon) + 
    labs(x = "Amplicon", fill = "Species Comparison") + 
        theme_bw() + 
    theme(legend.position = "bottom")
```

********************************************************************************

#### Species level comparison
While the overall pairwise distance is greater between species than within species for the _Paenibacillus_ genus, it is important to understand how the within and between species pairwise distances compare for individual species. 
The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)). 
Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm. 
In general though for species levels classification the maximum within species distance should be less than the minimum between species distance. 
For example as the maximum within species pairwise distance for _P. marcerans_ is 0.06 and the minimum between species pairwise distance for _P. marcerans_ and _P. barengoltzii_ is 0.02 (Fig. \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two species is not possible.

```{r distHeatMap, fig.cap = "Pairwise distances between _Paenibacillus_ species (A) V12 and (B) V4 amplicon regions. Fill color indicates the mean pairwise sequence distance within and between species. The text indicates the maximum pairwise distance for within species comparisons, values along the diagonal, and maximum pairwise distance for between species comparisons.", fig.height = 8}
v12_heat <- v12_dist_anno_df %>% 
    group_by(Keys_Species, Keys2_Species, group_comp) %>% 
    summarise(mean_dist = mean(distance),
              max_dist = max(distance),
              min_dist = min(distance)) %>% 
    mutate(lab_dist = if_else(group_comp == "within", max_dist, min_dist)) %>% 
    ggplot() + 
    geom_raster(aes(x = Keys_Species, 
                    y = Keys2_Species, 
                    fill = mean_dist)) +
    geom_text(aes(x = Keys_Species, y = Keys2_Species, label = round(lab_dist,2)), color = "white") + 
    labs(x = "Species",y = "Species", fill = "Mean Dist") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))

v4_heat <- v4_dist_anno_df %>% 
    filter(Keys != "329842", Keys2 != "329842") %>% 
    group_by(Keys_Species, Keys2_Species, group_comp) %>% 
    summarise(mean_dist = mean(distance),
              max_dist = max(distance),
              min_dist = min(distance)) %>% 
    mutate(lab_dist = if_else(group_comp == "within", max_dist, min_dist)) %>% 
    ggplot() + 
    geom_raster(aes(x = Keys_Species, 
                    y = Keys2_Species, 
                    fill = mean_dist)) +
    geom_text(aes(x = Keys_Species, y = Keys2_Species, label = round(lab_dist,2)), color = "white") + 
    labs(x = "Species",y = "Species", fill = "Mean Dist") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))

ggpubr::ggarrange(v12_heat, v4_heat,labels = "AUTO",ncol = 1, nrow = 2,align = "v")
```

## Conclusion
Here we demonstrate how the `metagenomeFeatures` package in conjunction with one of the associated 16S rRNA database packages, `greengenes13.5MgDb`, and other R packages, can be used to evaluate whether species level taxonomic classification is possible for a specific amplicon region. 
The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`clas object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).

********************************************************************************

## Session Information
### System Information
```{r}
s_info <- devtools::session_info(include_base = FALSE)
pander::pander(s_info$platform)
```

### Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>% 
    knitr::kable(booktabs = TRUE)
```

### 2018-04-01 inst/manuscript/manuscript_supplemental.Rmd

@@
 We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus. 
-Taxonomic resolution can vary by clade and amplicon regions, though the extent to which it varies is not well characterized.
+While similar to determining whether a sequence represents a novel species, here we are only interested in determining whether the 16S rRNA region of interest contains sufficient information for species level taxonomic assignment.
+Taxonomic resolution can vary by clade and amplicon regions, though the extent to which it varies has not been characterized.
 
-Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions
+Here we demonstrate how `metagenomeFeatures` and `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions
 Using the Greengenes 13.5, `greengenes13.5MgDb` we will first subset the database to obtain _Paenibacillus_ taxonomic and sequence data. 
@@
 ## Required Packages  
-Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix. 
-The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures.
+Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@decipher]. 
+The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse;@ggpubr].
 
@@
 We first subset the Greeengenes 13.5 database using the `mgDb_select` function. 
-Then summarize the taxonomy data using the `tidyverse` package. 
+Then summarize the taxonomy data using functions from `tidyverse` package, specifically `dplyr`, `stringr` and `forcats` functions for manipulating `data.frames`, `strings`, and `factor` vectors respectively.
 <!-- Loading metagenomeFeatues and greengenes db -->
@@
 ```{r}
-paeni_16S <- mgDb_select(gg13.5MgDb, 
+paeni_16S <- metagenomeFeatures::mgDb_select(gg13.5MgDb, 
                           type = c("taxa","seq"), 
@@
 
-Generating pairwise distance matrix for taxonomic resolution analysis and converting distance matrix to data frame for analysis.
+Generating pairwise distance matrix using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to data frame for analysis.
 

### 2018-04-02 R/gg13.8_97MgDb.R
using gg13.8 85 otus instead of 97

@@ -1,26 +1,0 @@
-#' MgDb-class object with Greengenes Database Version 13.8 97\% OTUs
-#'
-#' Example \link[=MgDb]{MgDb-class} object with Greengenes Database Version
-#' 13.8 97\% OTUs.
-#' @return MgDb-class object
-#' @export
-#' @examples
-#' get_gg13.8_97MgDb()
-get_gg13.8_97MgDb <- function(){
-    ## Source data generated using the make_greengenes13.8_97.R script in inst/scripts
-
-    metadata_file <- system.file("extdata", 'gg13.8_97_metadata.RData',
-                           package = "metagenomeFeatures")
-    load(metadata_file)
-
-    gg_db_file <- system.file("extdata", 'gg13.8_97.sqlite',
-                                package = "metagenomeFeatures")
-
-    gg_tree_file <- system.file("extdata", "gg13.8_97.tre",
-                                  package = "metagenomeFeatures")
-
-    ## Creating a new MgDb class object with gg13.8_97 data
-    newMgDb(db_file = gg_db_file,
-            tree = gg_tree_file,
-            metadata =  metadata)
-}

### 2018-04-02 inst/scripts/make_greengenes13.8_85.R

@@ -1,0 +1,80 @@
+### Code to generate source files for a MgDb-object with data from the
+### Greengenes 13.8 database 97% OTUs
+library(DECIPHER)
+library(Biostrings)
+library(metagenomeFeatures) ## Needed for the make_mgdb_sqlite function
+
+## Database URL
+db_root_url <- "ftp://greengenes.microbio.me/greengenes_release/gg_13_8_otus/"
+taxa_url <- paste0(db_root_url, "taxonomy/85_otu_taxonomy.txt")
+aligned_seq_url <- paste0(db_root_url, "rep_set_aligned/85_otus.fasta")
+seq_url <- paste0(db_root_url, "rep_set/85_otus.fasta")
+tree_url <- paste0(db_root_url, "trees/85_otus_unannotated.tree")
+
+## Downloaded files
+taxa_file <- tempfile()
+seq_file <- tempfile()
+tree_file <- "../extdata/gg13.8_85.tre"
+
+## MD5 check sums from initial download
+taxa_md5 <- "a7f5343339c75036ae40c7a714c87eb"
+seq_md5 <- "6c571af57ab0f72af9288b72f6c50206"
+tree_md5 <- "a40fbc1d50efa59b77ba59ae226d90d2"
+
+## MgDb database files name
+db_file <- "../extdata/gg13.8_85.sqlite"
+metadata_file <- "../extdata/gg13.8_85_metadata.RData"
+
+### Download database files ####################################################
+download_db <- function(url, file_name, md5){
+    ## Downloade file and check to make sure MD5 checksum matches checksum for
+    ## previously downloaded version
+
+    download.file(url,file_name)
+    new_md5 <- tools::md5sum(file_name)
+    if (md5 != new_md5) warning("checksum does not match downloaded file.")
+}
+
+## Taxa Data
+download_db(taxa_url, taxa_file, taxa_md5)
+
+## Seq Data
+download_db(seq_url, seq_file, seq_md5)
+
+## Tree Data
+download_db(tree_url, tree_file, tree_md5)
+
+### Create SQLite DB with Taxa and Seq Data ####################################
+### Parse greengenes taxonomy
+parse_greengenes <- function(taxonomy_file){
+    taxa <- read.delim(taxonomy_file, stringsAsFactors = FALSE, header = FALSE)
+    keys <- taxa[,1]
+    taxa <- strsplit(taxa[,2],split = "; ")
+    taxa <- t(sapply(taxa,function(i){i}))
[…]

### 2018-04-02 tests/testthat/test_mgDb-select.R

@@ -123,10 +123,10 @@ test_that("MgDb-class select all",{
 
 })
 
-## Using Entero GG97 as bug first identified when developing
+## Using Entero GG85 as bug first identified when developing
 ## database-explore.Rmd vignette
 test_that("MgDb-class select equal entries",{
-    test_select_entero <- mgDb_select(get_gg13.8_97MgDb(),
+    test_select_entero <- mgDb_select(get_gg13.8_85MgDb(),
                                       type = "all",
                                       keys = "Enterobacteriaceae",
                                       keytype = "Family")

### 2018-04-02 R/gg13.8_85MgDb.R
using gg13.8 85 otus instead of 97

@@ -1,0 +1,26 @@
+#' MgDb-class object with Greengenes Database Version 13.8 97\% OTUs
+#'
+#' Example \link[=MgDb]{MgDb-class} object with Greengenes Database Version
+#' 13.8 97\% OTUs.
+#' @return MgDb-class object
+#' @export
+#' @examples
+#' get_gg13.8_97MgDb()
+get_gg13.8_85MgDb <- function(){
+    ## Source data generated using the make_greengenes13.8_85.R script in inst/scripts
+
+    metadata_file <- system.file("extdata", 'gg13.8_85_metadata.RData',
+                           package = "metagenomeFeatures")
+    load(metadata_file)
+
+    gg_db_file <- system.file("extdata", 'gg13.8_85.sqlite',
+                                package = "metagenomeFeatures")
+
+    gg_tree_file <- system.file("extdata", "gg13.8_85.tre",
+                                  package = "metagenomeFeatures")
+
+    ## Creating a new MgDb class object with gg13.8_85 data
+    newMgDb(db_file = gg_db_file,
+            tree = gg_tree_file,
+            metadata =  metadata)
+}

### 2018-04-02 R/mgDb-class.R

@@ -28,7 +28,7 @@ setOldClass(c("tbl_dbi"))
 #' The MgDb-class object contains sequence, taxonomic data, and a phylogenetic
 #' tree (optional) for a 16S rRNA taxonomic database, see the
 #' \pkg{greengenes13.5MgDb} package as an example database. The
-#' \code{get_gg13.8_97MgDb()} function in \pkg{metagenomeFeatures} exports a small
+#' \code{get_gg13.8_85MgDb()} function in \pkg{metagenomeFeatures} exports a small
 #' subset of the database in the \pkg{greengenes13.5MgDb} annotation package as
 #' an example of a MgDb-class object.
 #' @aliases mgdb
@@ -38,8 +38,8 @@ setOldClass(c("tbl_dbi"))
 #' @slot metadata associated metadata for the database
 #' @export
 #' @examples
-#' # example MgDb-class object, Greengenes 13.8 97% OTUs database.
-#' gg97 <- get_gg13.8_97MgDb()
+#' # example MgDb-class object, Greengenes 13.8 85% OTUs database.
+#' gg85 <- get_gg13.8_85MgDb()
 #' @note Currently the only database with a MgDb package is the
 #'   \href{http://greengenes.secondgenome.com/}{Greengenes database} (version
 #'   13.5), additional packages are planned.
@@ -170,17 +170,17 @@ make_mgdb_sqlite <- function(db_name, db_file, taxa_tbl, seqs) {
 #' @export
 #'
 #' @examples
-#' metadata_file <- system.file("extdata", 'gg13.8_97_metadata.RData',
+#' metadata_file <- system.file("extdata", 'gg13.8_85_metadata.RData',
 #'     package = "metagenomeFeatures")
 #' load(metadata_file)
 #'
-#' gg_db_file <- system.file("extdata", 'gg13.8_97.sqlite',
+#' gg_db_file <- system.file("extdata", 'gg13.8_85.sqlite',
 #'                           package = "metagenomeFeatures")
 #'
-#' gg_tree_file <- system.file("extdata", "gg13.8_97.tre",
+#' gg_tree_file <- system.file("extdata", "gg13.8_85.tre",
 #'                             package = "metagenomeFeatures")
 #'
[…]

### 2018-04-02 R/mgDb-taxa_accessors.R

@@ -36,8 +36,8 @@
 #' @return tbl_df
 #'
 #' @examples
-#' gg97 <- get_gg13.8_97MgDb()
-#' taxa_keys(gg97, keytype = "Phylum")
+#' gg85 <- get_gg13.8_85MgDb()
+#' taxa_keys(gg85, keytype = "Phylum")
 #' @exportMethod taxa_keys
 #' @rdname taxa_keys
 setGeneric("taxa_keys", signature = "mgdb",
@@ -66,8 +66,8 @@ setMethod("taxa_keys", "MgDb",
 #' @return character vector
 #' @note Same function as \code{\link{taxa_keytypes}}.
 #' @examples
-#' gg97 <- get_gg13.8_97MgDb()
-#' taxa_columns(gg97)
+#' gg85 <- get_gg13.8_85MgDb()
+#' taxa_columns(gg85)
 #' @exportMethod taxa_columns
 #' @rdname taxa_columns
 setGeneric("taxa_columns", signature = "mgdb",
@@ -91,8 +91,8 @@ setMethod("taxa_columns", "MgDb",
 #' @param mgdb object of MgDB class
 #' @return tbl_df
 #' @examples
-#' gg97 <- get_gg13.8_97MgDb()
-#' taxa_keytypes(gg97)
+#' gg85 <- get_gg13.8_85MgDb()
+#' taxa_keytypes(gg85)
 #' @exportMethod taxa_keytypes
 #' @rdname taxa_keytypes
 setGeneric("taxa_keytypes", signature = "mgdb",

### 2018-04-02 R/mgDb_method_annotateFeatures.R

@@ -69,10 +69,10 @@
 #' @return mgFeatures-class object
 #' @examples
 #' ## MgDb with mock community ids
-#' gg97 <- get_gg13.8_97MgDb()
+#' gg85 <- get_gg13.8_85MgDb()
 #' ## generating mgFeatures object
 #' data(mock_query_df)
-#' mock_mgF <- annotateFeatures(gg97, mock_query_df)
+#' mock_mgF <- annotateFeatures(gg85, mock_query_df)
 #'
 #' @rdname annotateFeatures-MgDb-method
 setGeneric("annotateFeatures", signature = "mgdb",

### 2018-04-02 R/mgDb_method_select.R

@@ -130,19 +130,19 @@
 #'   tree of class phylo; 'all' - list with the dataframe, DNAStringSet, and
 #'   phylo.
 #' @examples
-#' gg97 <- get_gg13.8_97MgDb()
+#' gg97 <- get_gg13.8_85MgDb()
 #' # select taxa only
-#' mgDb_select(gg97, type = "taxa",
+#' mgDb_select(gg85, type = "taxa",
 #'      keys = c("Vibrionaceae", "Enterobacteriaceae"),
 #'      keytype = "Family")
 #'
 #' # select seq only
-#'  mgDb_select(gg97, type = "seq",
+#'  mgDb_select(gg85, type = "seq",
 #'       keys = c("Vibrionaceae", "Enterobacteriaceae"),
 #'       keytype = "Family")
 #'
 #' # select all taxa, seq, and tree
-#' mgDb_select(gg97, type = "all",
+#' mgDb_select(gg85, type = "all",
 #'        keys = c("Vibrionaceae", "Enterobacteriaceae"),
 #'        keytype = "Family")
 #' @rdname select-MgDb-method

### 2018-04-02 R/mgDb-class.R
revised documentation structure

@@ -30,7 +30,7 @@ setOldClass(c("tbl_dbi"))
 #' \pkg{greengenes13.5MgDb} package as an example database. The
 #' \code{get_gg13.8_85MgDb()} function in \pkg{metagenomeFeatures} exports a small
 #' subset of the database in the \pkg{greengenes13.5MgDb} annotation package as
-#' an example of a MgDb-class object.
+#' an example MgDb-class object.
 #' @aliases mgdb
 #' @slot seq database reference sequences
 #' @slot tree reference phylogenetic tree
@@ -71,7 +71,7 @@ setClass("MgDb",
 #'
 #' @examples
 #' \dontrun{
-#' make_mgdb_sqlite(db_name = "greengenes13.8_97",
+#' make_mgdb_sqlite(db_name = "greengenes13.8_85",
 #'     db_file = db_file,
 #'     taxa_tbl = taxa_tbl,
 #'     seqs = seqs)
@@ -280,62 +280,51 @@ setMethod("show", "MgDb",
 
 ## Accessors -------------------------------------------------------------------
 
-#' MgDb tree slot accessor
+#' MgDb-class accessors
 #'
-#' @param mgdb  MgDb class object
+#' Accessors for \linkS4class{MgDb}-class object slots. \code{mgDb_seq} - sequence slot, \code{mgDb_taxa} - taxa slot, \code{mgDb_tree} - phylogenetic tree slot, and \code{mgDb_meta} - metadata slot.
 #'
-#' @return phylo class object
-#' @export
+#' @name mgDb_
+#' @param mgdb MgDb-class object.
 #'
+#' @return appropriate class object for the slot accessed
 #' @examples
 #' gg85 <- get_gg13.8_85MgDb()
-#' mgdb_tree(gg85)
-mgdb_tree <- function(mgdb){
+#' mgDb_seq(gg85)
+#' mgDb_taxa(gg85)
+#' mgDb_tree(gg85)
+#' mgDb_meta(gg85)
+NULL
+
+
+# MgDb tree slot accessor
+#' @export
+#' @rdname mgDb_
+mgDb_tree <- function(mgdb){
     ## Add assertion for MgDb class object
     mgdb@tree
 }
 
-#' MgDb seq slot accessor
-#'
-#' @param mgdb  MgDb class object
-#'
-#' @return DNAStringSet class object
[…]

### 2018-04-02 R/mgDb-taxa_accessors.R

@@ -4,6 +4,22 @@
 ##
 ### ============================================================================
 
+#' MgDb-class Taxa slot helper functions
+#'
+#' Helper functions for for \linkS4class{MgDb}-class taxa slot. \code{taxa_columns} - taxa slot column names, \code{taxa_keytypes} - taxa slot keytypes (values used with \code{taxa_columns} and \code{mgDb_select} functions), and \code{taxa_keys} - database values for a specific keytype.
+#'
+#' @name taxa_
+#' @param mgdb MgDb-class object.
+#'
+#' @return appropriate class object for the slot accessed
+#' @examples
+#' gg85 <- get_gg13.8_85MgDb()
+#' taxa_columns(gg85)
+#' taxa_keytypes(gg85)
+#' taxa_keys(gg85, keytype = "Phylum")
+NULL
+
+
 ### Taxa keys function ---------------------------------------------------------
 .taxa_keys <- function(mgdb, keytype){
     if (length(keytype) > 0) {
@@ -28,23 +44,14 @@
     }
 }
 
-#' Taxonomy values for a given keytype
-#'
-#' @name taxa_keys
-#' @param mgdb object of MgDB class
-#' @param keytype taxonomic classification level
-#' @return tbl_df
-#'
-#' @examples
-#' gg85 <- get_gg13.8_85MgDb()
-#' taxa_keys(gg85, keytype = "Phylum")
-#' @exportMethod taxa_keys
-#' @rdname taxa_keys
+# Taxonomy values for a given keytype
+#' @export
+#' @rdname taxa_
 setGeneric("taxa_keys", signature = "mgdb",
            function(mgdb, keytype) standardGeneric("taxa_keys"))
 
 ## MgDb taxa_keys method
-#' @rdname taxa_keys
+#' @rdname taxa_
 #' @aliases taxa_keys,MgDb-method
 setMethod("taxa_keys", "MgDb",
           function(mgdb, keytype) .taxa_keys(mgdb, keytype))
@@ -60,22 +67,14 @@ setMethod("taxa_keys", "MgDb",
 }
 
 #' Column names for MgDb taxonomy slot object
-#'
-#' @name taxa_columns
-#' @param mgdb object of MgDB class
-#' @return character vector
-#' @note Same function as \code{\link{taxa_keytypes}}.
-#' @examples
[…]

### 2018-04-02 R/mock_mgF.R

@@ -12,6 +12,9 @@
 #'
 #' @format mgFeatures
 #'
+#' @source
+#' \url{https://qiita.ucsd.edu}
+#'
 #' @examples
 #' data(mock_mgF)
 "mock_mgF"

### 2018-04-02 R/mock_query_df.R

@@ -13,7 +13,7 @@
 #' @format data.frame
 #'
 #' @source
-#' \url{http://bioconductor.org/packages/release/data/experiment/html/msd16s.html}
+#' \url{https://qiita.ucsd.edu}
 #'
 #' @examples
 #' data(mock_query_df)

### 2018-04-02 R/gg13.8_85MgDb.R
fixing long code lines

@@ -7,7 +7,7 @@
 #' @examples
 #' get_gg13.8_85MgDb()
 get_gg13.8_85MgDb <- function(){
-    ## Source data generated using the make_greengenes13.8_85.R script in inst/scripts
+    ## Data generated using make_greengenes13.8_85.R script in inst/scripts
 
     metadata_file <- system.file("extdata", 'gg13.8_85_metadata.RData',
                            package = "metagenomeFeatures")

### 2018-04-02 R/mgDb-class.R

@@ -28,9 +28,9 @@ setOldClass(c("tbl_dbi"))
 #' The MgDb-class object contains sequence, taxonomic data, and a phylogenetic
 #' tree (optional) for a 16S rRNA taxonomic database, see the
 #' \pkg{greengenes13.5MgDb} package as an example database. The
-#' \code{get_gg13.8_85MgDb()} function in \pkg{metagenomeFeatures} exports a small
-#' subset of the database in the \pkg{greengenes13.5MgDb} annotation package as
-#' an example MgDb-class object.
+#' \code{get_gg13.8_85MgDb()} function in \pkg{metagenomeFeatures} exports a
+#' small subset of the database in the \pkg{greengenes13.5MgDb} annotation
+#' package as an example MgDb-class object.
 #' @aliases mgdb
 #' @slot seq database reference sequences
 #' @slot tree reference phylogenetic tree
@@ -186,8 +186,8 @@ make_mgdb_sqlite <- function(db_name, db_file, taxa_tbl, seqs) {
 #'         metadata =  metadata)
 #'
 newMgDb <- function(db_file, tree, metadata){
-    ## db_file is character string, file exists, is a properly formatted sqlite database
-    ## Check tree is either a phylo class object or newick tree file
+    ## db_file is character string, file exists, is a properly formatted sqlite
+    ## database Check tree is either a phylo class object or newick tree file
     ## Check metadata is list with required entries
 
     ## sequence slot
@@ -282,7 +282,9 @@ setMethod("show", "MgDb",
 
 #' MgDb-class accessors
 #'
-#' Accessors for \linkS4class{MgDb}-class object slots. \code{mgDb_seq} - sequence slot, \code{mgDb_taxa} - taxa slot, \code{mgDb_tree} - phylogenetic tree slot, and \code{mgDb_meta} - metadata slot.
+#' Accessors for \linkS4class{MgDb}-class object slots. \code{mgDb_seq} -
+#' sequence slot, \code{mgDb_taxa} - taxa slot, \code{mgDb_tree} - phylogenetic
+#' tree slot, and \code{mgDb_meta} - metadata slot.
 #'
 #' @name mgDb_
 #' @param mgdb MgDb-class object.

### 2018-04-02 R/mgDb-taxa_accessors.R

@@ -6,7 +6,10 @@
 
 #' MgDb-class Taxa slot helper functions
 #'
-#' Helper functions for for \linkS4class{MgDb}-class taxa slot. \code{taxa_columns} - taxa slot column names, \code{taxa_keytypes} - taxa slot keytypes (values used with \code{taxa_columns} and \code{mgDb_select} functions), and \code{taxa_keys} - database values for a specific keytype.
+#' Helper functions for for \linkS4class{MgDb}-class taxa slot.
+#' \code{taxa_columns} - taxa slot column names, \code{taxa_keytypes} - taxa
+#' slot keytypes (values used with \code{taxa_columns} and \code{mgDb_select}
+#' functions), and \code{taxa_keys} - database values for a specific keytype.
 #'
 #' @name taxa_
 #' @param mgdb MgDb-class object.

### 2018-04-02 R/mgDb_method_annotateFeatures.R

@@ -18,7 +18,8 @@
 
     if (is.data.frame(query)) {
         if (is.element("Keys", colnames(query))) {
-            query$Keys <- as.character(query$Keys) # process Keys column as character
+            # process Keys column as character
+            query$Keys <- as.character(query$Keys)
             select_keys <- query$Keys
         } else {
             stop("Need 'Keys' column in 'query' with database ids")
@@ -26,7 +27,7 @@
     } else if (is.vector(query)) { # db_keys used
         select_keys <- as.character(query)
     } else {## query not a data frame or vector
-        stop("query not a vector or data.frame, see documentation for query requirements")
+        stop("query must be a vector or data.frame")
     }
 
 

### 2018-04-02 R/mgDb_method_select.R

@@ -25,7 +25,9 @@
     # selecting desired rows
     if (!is.null(keys)) {
         if (keytype !=  "Keys") {
-            level_id <- stringr::str_sub(string = keytype, start = 1,end = 1) %>%
+            level_id <- stringr::str_sub(string = keytype,
+                                         start = 1,
+                                         end = 1) %>%
                 tolower() %>% rep(length(keys))
             if (metaDb$DB_TYPE_NAME == "GreenGenes") {
                 keys <- stringr::str_c(level_id,keys,sep = "__")
@@ -55,7 +57,8 @@
 
 .select.tree <- function(tree, ids){
     drop_tips <- tree$tip.label[!(tree$tip.label %in% ids)]
-    # drop.tip return class phy defining class to match mgFeature class description
+    # drop.tip return class phy defining
+    # class to match mgFeature class description
     ape::drop.tip(tree,drop_tips) %>% ape::as.phylo()
 }
 
@@ -89,11 +92,13 @@
     # vector with all objects to return
     if ("taxa" %in% type || type == "all") {
         ## Removing decipher columns
-        select_obj$taxa <- dplyr::select(taxa_df, -row_names, -description, -identifier)
+        select_obj$taxa <- dplyr::select(taxa_df, -row_names,
+                                         -description, -identifier)
     }
 
     if ("seq" %in% type || type == "all") {
-        select_obj$seq <- .select.seq(mgDb_seq(mgdb), taxa_df$row_names, taxa_df$Keys)
+        select_obj$seq <- .select.seq(mgDb_seq(mgdb),
+                                      taxa_df$row_names, taxa_df$Keys)
     }
 
     if ("tree" %in% type || type == "all") {

### 2018-04-02 R/mgFeatures-class.R

@@ -102,7 +102,9 @@ setMethod("[", "mgFeatures",
 
 #' mgFeatures accessors
 #'
-#' Accessors for \linkS4class{mgFeatures}-class object slots. \code{mgF_seq} - refDbSeq slot, \code{mgF_taxa} - taxa slot, \code{mgF_tree} - phylogenetic tree slot, and \code{mgF_meta} - metadata slot.
+#' Accessors for \linkS4class{mgFeatures}-class object slots. \code{mgF_seq} -
+#' refDbSeq slot, \code{mgF_taxa} - taxa slot, \code{mgF_tree} - phylogenetic
+#' tree slot, and \code{mgF_meta} - metadata slot.
 #'
 #' @name mgF_
 #' @param mgF mgFeatures-class object.

### 2018-04-02 R/mock_mgF.R

@@ -8,7 +8,8 @@
 #' (https://qiita.ucsd.edu).
 #'
 #' Bokulich, Nicholas A., et al. "Quality-filtering vastly improves diversity
-#' estimates from Illumina amplicon sequencing." Nature methods 10.1 (2013): 57-59.
+#' estimates from Illumina amplicon sequencing." Nature methods 10.1 (2013):
+#' 57-59.
 #'
 #' @format mgFeatures
 #'

### 2018-04-02 R/mock_query_df.R

@@ -1,14 +1,15 @@
 #' Example Query Data Frame
 #'
 #' Example \code{query_df} for use in generating a \link{mgFeatures-class}
-#' object using the \link{annotateFeatures} \link[=MgDb]{MgDb-class} method.
-#' The dataset contains the OTU ids and Greengenes database version 13.5 ids.
+#' object using the \link{annotateFeatures} \link[=MgDb]{MgDb-class} method. The
+#' dataset contains the OTU ids and Greengenes database version 13.5 ids.
 #' Specifically, the OTU centers came from a mock community dataset from the
 #' study Bokulich et al. 2013, the OTU ids were extracted from a biom file
 #' downloaded from QIITA (https://qiita.ucsd.edu).
 #'
 #' Bokulich, Nicholas A., et al. "Quality-filtering vastly improves diversity
-#' estimates from Illumina amplicon sequencing." Nature methods 10.1 (2013): 57-59.
+#' estimates from Illumina amplicon sequencing." Nature methods 10.1 (2013):
+#' 57-59.
 #'
 #' @format data.frame
 #'

### 2018-04-02 R/mgDb-class.R
fixed function names

@@ -261,17 +261,17 @@ setMethod("show", "MgDb",
           function(object){
               cat(class(object), "object:")
               print("Metadata")
-              metadata <- mgdb_meta(object)
+              metadata <- mgDb_meta(object)
               for (i in names(metadata)) {
                   cat("|", i, ": ", metadata[[i]], "\n", sep = "")
               }
               print("Sequence Data:")
               print("DECIPHER formatted seqDB")
               print("Taxonomy Data:")
-              print(mgdb_taxa(object))
+              print(mgDb_taxa(object))
               print("Tree Data:")
-              if (!is.null(mgdb_tree(object))) {
-                  ape::print.phylo(mgdb_tree(object))
+              if (!is.null(mgDb_tree(object))) {
+                  ape::print.phylo(mgDb_tree(object))
               } else {
                   print("Tree not available")
               }

### 2018-04-02 R/mgDb_method_select.R
more long line fixes

@@ -68,7 +68,8 @@
     select_types <- c("seq","taxa", "tree", "all")
     if (FALSE %in% (type %in% select_types)) {
         bad_type <- type[!(type %in% select_types)]
-        stop(paste(bad_type, "not valid type value, type must be either 'seq', 'taxa', 'tree', 'all' or a character vector with types"))
+        stop(paste(bad_type, "not valid type value, type must be either 'seq',",
+                    "'taxa', 'tree', 'all' or a character vector with types"))
     }
 
     if (is.null(keys) != is.null(keytype)) {

### 2018-04-02 R/mgFeatures-class.R

@@ -76,7 +76,8 @@ mgFeatures <- function(taxa, tree = NULL, seq = NULL, metadata){
 ## subset ----------------------------------------------------------------------
 .subset_tree <- function(tree, ids){
     drop_tips <- tree$tip.label[!(tree$tip.label %in% ids)]
-    # drop.tip return class phy defining class to match mgFeature class description
+    # drop.tip return class phy defining class to match mgFeature class
+    # description
     ape::drop.tip(tree,drop_tips) # %>% ape::as.phylo()
 }
 

### 2018-09-25 R/qiita_study_94_gg_ids.R
convert qiita data to rda & move to data folder

@@ -1,0 +1,7 @@
+#' QIITA Dataset
+#' For this vignette we are using 16S rRNA data from Rousk et al. 2010, a soil microbiome study, https://qiita.ucsd.edu/study/description/94.
+#' A BIOM and qiime mapping file for the study can be obtained from QIITA.
+#' A vector of Greengenes for the study cluster centers is included in this package for use in this vignette.
+#'
+#' data(qiita_study_94_gg_ids)
+"qiita_study_94_gg_ids"

### 2018-09-25 R/mgDb-class.R
update source documentation & vignette

@@ -43,6 +43,7 @@ setOldClass(c("tbl_dbi"))
 #' @note Currently the only database with a MgDb package is the
 #'   \href{http://greengenes.secondgenome.com/}{Greengenes database} (version
 #'   13.5), additional packages are planned.
+#' @name MgDb-class
 #' @rdname MgDb-class
 #' @return MgDb-class object
 #' @importFrom dbplyr tbl_sql
@@ -256,6 +257,12 @@ setValidity("MgDb", function(object) {
 #' Display summary of MgDb-class object
 #' @param object MgDb-class object
 #' @return MgDb-class summary
+#' @name MgDb-methods
+#' @rdname MgDb-methods
+#' @aliases show,MgDb-method
+#' @examples
+#' gg85 <- get_gg13.8_85MgDb()
+#' show(gg85)
 #' @export
 setMethod("show", "MgDb",
           function(object){
@@ -286,7 +293,8 @@ setMethod("show", "MgDb",
 #' sequence slot, \code{mgDb_taxa} - taxa slot, \code{mgDb_tree} - phylogenetic
 #' tree slot, and \code{mgDb_meta} - metadata slot.
 #'
-#' @name mgDb_
+#' @name MgDb-accessor
+#' @rdname MgDb-accessor
 #' @param mgdb MgDb-class object.
 #'
 #' @return appropriate class object for the slot accessed
@@ -301,7 +309,8 @@ NULL
 
 # MgDb tree slot accessor
 #' @export
-#' @rdname mgDb_
+#' @name MgDb-accessor
+#' @rdname MgDb-accessor
 mgDb_tree <- function(mgdb){
     ## Add assertion for MgDb class object
     mgdb@tree
@@ -309,7 +318,8 @@ mgDb_tree <- function(mgdb){
 
 # MgDb seq slot accessor
 #' @export
-#' @rdname mgDb_
+#' @name MgDb-accessor
+#' @rdname MgDb-accessor
 mgDb_seq <- function(mgdb){
     ## Add assertion for MgDb class object
     mgdb@seq
@@ -317,7 +327,8 @@ mgDb_seq <- function(mgdb){
 
 # MgDb taxa slot accessor
 #' @export
-#' @rdname mgDb_
+#' @name MgDb-accessor
+#' @rdname MgDb-accessor
 mgDb_taxa <- function(mgdb){
     ## Add assertion for MgDb class object
     mgdb@taxa
[…]

### 2018-09-25 R/mgFeatures-class.R

@@ -1,21 +1,19 @@
 ## ========================= mgFeature Class ========================
-#' mgFeature-class object
+#' Class mgFeature
 #'
-#' Object contains taxonomic annotation and reference sequence data for
+#' Class defines taxonomic annotation and reference sequence data for
 #' classified OTUs. The class extends the \link[S4Vectors]{DataFrame}
 #' class with a slot with a subset of the reference database sequences and
 #' phylogenetic tree for taxonomically classified OTUs, along with an additional
 #' slot for metadata including information on the database source.
 #'
-#' @slot metadata list
-#' @slot refDbSeq DNAStringSetOrNull
-#' @slot refDbTree phyloOrNULL
 #' @return mgFeature class object
-#' @export
-#' @examples
-#' data(mock_mgF)
+#' @name mgFeatures-class
 #' @rdname mgFeatures-class
 #' @importFrom S4Vectors DataFrame
+#' @exportClass mgFeatures
+#' @examples
+#' data(mock_mgF)
 setClass("mgFeatures",
          slots = list(metadata = "list",
                       refDbSeq = "DNAStringSetOrNull",
@@ -44,18 +42,18 @@ setValidity("mgFeatures", function(object) {
 
 
 #' mgFeatures-class constructor
-#'
 #' @param taxa a DataFrame-class or object that can be coerced into a DataFrame
 #' @param tree a phylo-class object with phylogenetic tree
 #' @param metadata a list
 #' @param seq DNAStringSet-object with feature sequences
-#'
 #' @return mgFeatures-class object
+#' @name mgFeatures-class
+#' @rdname mgFeatures-class
 #' @export
 #'
 #' @examples
 #' mgFeatures(taxa = data.frame(), metadata = list())
-mgFeatures <- function(taxa, tree = NULL, seq = NULL, metadata){
+mgFeatures <- function(taxa = data.frame(), tree = NULL, seq = NULL, metadata){
     new("mgFeatures",
         DataFrame(taxa),
         metadata = metadata,
@@ -81,6 +79,14 @@ mgFeatures <- function(taxa, tree = NULL, seq = NULL, metadata){
     ape::drop.tip(tree,drop_tips) # %>% ape::as.phylo()
 }
 
+#' mgFeatures-class subset method
[…]

### 2018-10-18 R/DNAStringSetOrNull-class.R
update imports, documentation

@@ -8,4 +8,6 @@ setOldClass("DNAStringSetOrNull")
 # the component data classes is not available, in which case NULL will be used
 # instead.
 #' @keywords internal
+#' @importClassesFrom Biostrings DNAStringSet
+#' @importClassesFrom Biostrings QualityScaledDNAStringSet
 setClassUnion("DNAStringSetOrNull", c("DNAStringSet", "NULL"))

### 2018-10-18 R/mgDb-taxa_accessors.R

@@ -25,6 +25,11 @@ NULL
 
 
 ### Taxa keys function ---------------------------------------------------------
+#' Taxa keys functions
+#' import dplyr
+#' @param mgdb mgDb class
+#' @param keytype taxa keytype
+#' @keywords internals
 .taxa_keys <- function(mgdb, keytype){
     if (length(keytype) > 0) {
         nonvalid_keytype <- c()
@@ -40,8 +45,8 @@ NULL
             stop(msg)
         }
         mgDb_taxa(mgdb) %>%
-            dplyr::select_(keytype) %>%
-            dplyr::collect() %>%
+            select_(keytype) %>%
+            collect() %>%
             return()
     }else{
         mgDb_taxa(mgdb)
@@ -63,6 +68,9 @@ setMethod("taxa_keys", "MgDb",
 
 
 ### Taxa columns function ------------------------------------------------------
+#' taxa columns
+#' @param mgdb mdgdb object
+#' @keywords internals
 .taxa_columns = function(mgdb){
     taxa_cols <- colnames(mgDb_taxa(mgdb))
     ## Excluding decipher column names
@@ -84,6 +92,7 @@ setMethod("taxa_columns", "MgDb",
 
 
 ### taxa keytypes function -----------------------------------------------------
+#' @keywords internals
 .taxa_keytypes = function(mgdb){
     colnames(mgDb_taxa(mgdb))
 }

### 2018-10-18 R/mgDb_method_annotateFeatures.R

@@ -5,9 +5,12 @@
 ################################################################################
 ## mgDb_annotateFeatures -------------------------------------------------------
 
-## Generates an mgFeatures object when passed a vector of keys or dataframe
-## (e.g. query)
-
+#' Generates an mgFeatures object when passed a vector of keys or dataframe
+#' (e.g. query)
+#' @param mgdb mdgdb object
+#' @param query search param
+#' @importFrom dplyr right_join
+#' @keywords internals
 .mgDb_annotateFeatures <- function(mgdb, query) {
     ##### SELECT_KEYS
     ## check query type: vector or data.frame
@@ -38,7 +41,7 @@
 
     ##### ANNOTATED_DB
     if (is.data.frame(query)) {
-        annotated_db <- dplyr::right_join(query, filtered_db$taxa)
+        annotated_db <- right_join(query, filtered_db$taxa)
     } else {
         annotated_db <- as.data.frame(filtered_db$taxa)
     }

### 2018-10-18 R/mgDb_method_select.R

@@ -5,43 +5,62 @@
 ################################################################################
 ## Select ----------------------------------------------------------------------
 
+#' select sequences
+#' @param seqObj seq object
+#' @param row_names row names to select
+#' @param Keys keys to filter
+#' @importClassesFrom Biostrings DNAStringSet
+#' @importFrom Biostrings DNAStringSet
+#' @importFrom RSQLite dbGetQuery
+#' @importFrom DECIPHER Codec
+#' @keywords internals
 .select.seq <- function(seqObj, row_names, Keys){
-    result <- RSQLite::dbGetQuery(seqObj, paste("select row_names, sequence from _Seqs where row_names in ", paste0("(", paste0(row_names, collapse = ','), ")")))
-    seqs <- DECIPHER::Codec(result$sequence)
-    dnaStringSet <- Biostrings::DNAStringSet(seqs)
+    result <- dbGetQuery(seqObj, paste("select row_names, sequence from _Seqs where row_names in ", paste0("(", paste0(row_names, collapse = ','), ")")))
+    seqs <- Codec(result$sequence)
+    dnaStringSet <- DNAStringSet(seqs)
     names(dnaStringSet) <- Keys[match(result[,1], row_names)]
 
     dnaStringSet
 }
 
-
+#' select taxa
+#' @param taxaDb taxa object
+#' @param metaDb metadata object
+#' @param keytype keytypes to select
+#' @param keys keys to filter
+#' @param columns columns to select, defaults to all
+#' @import dplyr
+#' @importFrom stringr str_c
+#' @importFrom stringr str_sub
+#' @import lazyeval
+#' @keywords internals
 .select.taxa <- function(taxaDb, metaDb, keys, keytype, columns="all"){
     ## setting values when keys and keytypes are not defined
     if (is.null(keys)) {
-        keys <- taxaDb %>% dplyr::select(Keys) %>% dplyr::collect() %>% .$Keys
+        keys <- taxaDb %>% select(Keys) %>% collect() %>% .$Keys
         keytype <- "Keys"
     }
 
     # selecting desired rows
     if (!is.null(keys)) {
         if (keytype !=  "Keys") {
-            level_id <- stringr::str_sub(string = keytype,
+            level_id <- str_sub(string = keytype,
                                          start = 1,
                                          end = 1) %>%
                 tolower() %>% rep(length(keys))
             if (metaDb$DB_TYPE_NAME == "GreenGenes") {
-                keys <- stringr::str_c(level_id,keys,sep = "__")
[…]

### 2018-10-18 R/mgFeatures-class.R

@@ -72,11 +72,12 @@ mgFeatures <- function(taxa = data.frame(), tree = NULL, seq = NULL, metadata){
 ################################################################################
 
 ## subset ----------------------------------------------------------------------
+#' @importFrom ape drop.tip
 .subset_tree <- function(tree, ids){
     drop_tips <- tree$tip.label[!(tree$tip.label %in% ids)]
     # drop.tip return class phy defining class to match mgFeature class
     # description
-    ape::drop.tip(tree,drop_tips) # %>% ape::as.phylo()
+    drop.tip(tree,drop_tips) # %>% ape::as.phylo()
 }
 
 #' mgFeatures-class subset method

### 2018-10-18 R/utils.R

@@ -5,18 +5,6 @@ magrittr::`%>%`
 #' @import methods
 NULL
 
-#' @importClassesFrom  Biostrings DNAStringSet
-NULL
-
-#' @import Biobase
-NULL
-
-#' @import lazyeval
-NULL
-
-#' @import RSQLite
-NULL
-
 #' @import lattice
 NULL
 

### 2018-10-18 R/gg13.8_85MgDb.R
update imports, documentation

@@ -11,7 +11,7 @@ get_gg13.8_85MgDb <- function(){
 
     metadata_file <- system.file("extdata", 'gg13.8_85_metadata.RData',
                            package = "metagenomeFeatures")
-    load(metadata_file)
+    load(metadata_file, envir=globalenv())
 
     gg_db_file <- system.file("extdata", 'gg13.8_85.sqlite',
                                 package = "metagenomeFeatures")

### 2018-10-18 R/mgDb-class.R

@@ -7,7 +7,10 @@
 
 ## MgDb Class ------------------------------------------------------------------
 
-## Tree can be either rds with phylo class object or tree file
+#' Tree can be either rds with phylo class object or tree file
+#' @param tree_file location of tree
+#' @importFrom ape read.tree
+#' @keywords internals
 .load_tree <- function(tree_file){
     if (!file.exists(tree_file)) {
         stop(paste0("tree_file:", tree_file, " is not a valid file path"))
@@ -16,7 +19,7 @@
     if (grepl("rds",tree_file,ignore.case = TRUE)) {
         tree <- readRDS(tree_file)
     }else{
-        tree <- ape::read.tree(tree_file)
+        tree <- read.tree(tree_file)
     }
 
     tree
@@ -70,6 +73,9 @@ setClass("MgDb",
 #' @return writes SQLite file
 #' @keywords internal
 #'
+#' @importFrom Biostrings readDNAStringSet
+#' @import RSQLite
+#' @importFrom DECIPHER Seqs2DB
 #' @examples
 #' \dontrun{
 #' make_mgdb_sqlite(db_name = "greengenes13.8_85",
@@ -103,7 +109,7 @@ make_mgdb_sqlite <- function(db_name, db_file, taxa_tbl, seqs) {
     ## seqs either a fasta file or DNAStringSet
     if (is.character(seqs)) {
         if (file.exists(seqs)) {
-            seqs <- Biostrings::readDNAStringSet(seqs)
+            seqs <- readDNAStringSet(seqs)
         } else {
             stop("seqs is a character string but no file exists, check filename")
         }
@@ -136,15 +142,15 @@ make_mgdb_sqlite <- function(db_name, db_file, taxa_tbl, seqs) {
     rownames(taxa_tbl) <- 1:nrow(taxa_tbl)
 
     ## Create database with taxa and sequence data
-    db_conn <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
+    db_conn <- dbConnect(SQLite(), db_file)
 
-    DECIPHER::Seqs2DB(seqs = seqs, type = "DNAStringSet",
+    Seqs2DB(seqs = seqs, type = "DNAStringSet",
                       dbFile = db_conn, identifier = "MgDb")
 
     ### Adding taxonomic data to database
 
     # get seqs table from database
[…]

### 2018-11-07 inst/manuscript/supplemental_material.Rmd
combined database version

---
title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
subtitle: "_Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions."
author: "Nathan D. Olson and Nidhi Shah"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
---

# Background
16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance analysis.
A limitation to 16S rRNA amplicon sequencing is a lack of taxonomic resolution, where organisms are only identifiable to the genus or family level.
We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus. 
We are interested in determining whether the 16S rRNA region of interest contains sufficient information for species-level taxonomic assignment.
Taxonomic resolution varies by clade and amplicon regions. 
However, the extent to which taxonomic resolution varies is not well characterized.

Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions. 
Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s. 
_Paenibacillus_ spp. are facultative anaerobic bacteria present in a variety of environments including the soil, water, and can act as opportunistic pathogens in humans [@ouyang2008]. 
It has been shown that _Paenibacillus_ spp. will play an important role in sustainable agricultural industries [@grady2016]. 
Thus, an appropriate speciation of this genus is of an interest to the community.

We used V12 and V4 region as they represent two commonly used amplicons for 16S rRNA marker-gene surveys. 
We will use the Greengenes 13.5 database, accessed using the _greengenes13.5MgDb_, Silva 128 database, accessed using the _silva128.1MgDb_, and RDP 11.5 database, accessed using the _ribosomaldatabaseproject11.5MgDb_, as annotation packages for our analysis of the _Paenibacillus_ genus. 
First, we show the distribution of sequences belonging to _Paenibacillus_ in these three database, and then a way to combine different subsets of database sequences to build a custom database object. 
This helps investigators to pool sequences from different databases, to create an accurate and most representative database for their sample. 
This is especially helpful when a certain clade of interest is not very well represented in the databases.
We then evaluate 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ species for V12 and V4 regions. 
In many analyses, the Greengenes 13.5 database is used for demonstration purposes but the other `MgDb` annotation packages such as RDP 11.5, SILVA 128, can also be used.

# Required Packages 
In addition to _metagenomeFeatures_, _greengenes13.5MgDb_, _ribosomaldatabaseproject11.5MgDb_, and _silva128.1MgDb_, the _DECIPHER_, _tidyverse_, and _ggpubr_ packages are also used in the following analysis. 
Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@Wright2016-mo].
The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse;@ggpubr].

```{r message = FALSE, echo = TRUE}
library(tidyverse); packageVersion("tidyverse")
library(ggpubr); packageVersion("ggpubr")
library(gplots); packageVersion("gplots")
library(DECIPHER); packageVersion("DECIPHER")
library(metagenomeFeatures); packageVersion("metagenomeFeatures")
library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
# library(silva128.1MgDb, lib.loc = "~/Documents/Fall2018/installed_packages/"); packageVersion("silva128.1MgDb")
library(silva128.1MgDb); packageVersion("silva128.1MgDb")
library(ribosomaldatabaseproject11.5MgDb); packageVersion("ribosomaldatabaseproject11.5MgDb")
```

# Database Comparison
- Three databases
  - greengenes
  - silva
  - Ribosomal database project

- number of database entries and fraction of entries with RNAcentralIDs
- note that the mgDb_taxa slot is a tbl_dbi
```{r echo = FALSE}
get_rnacentral_entries <- function(mgdb){
   if("rnacentral_ids" %in% taxa_columns(mgdb)){
    rna_filt <- mgDb_taxa(mgdb) %>% 
      filter(!is.na(rnacentral_ids))
  } else {
    rna_filt <- mgDb_taxa(mgdb) %>% 
      filter(!is.na(RNAcentralID))    
  }
  rna_filt 
}

get_size_and_rnacentral <- function(mgdb){
  db_size <- mgDb_taxa(mgdb) %>% 
    tally() %>% collect() %>% as.numeric()

  db_wrna_id <- get_rnacentral_entries(mgdb) %>% 
    tally() %>% collect() %>% as.numeric()

  ## return size and size with rna central ids
  data_frame(n_entries = db_size, 
             n_entries_rnaid = db_wrna_id)
}

db_list <- list(gg = gg13.5MgDb, 
                slv = slv128.1MgDb, 
                rdp = rdp11.5MgDb)

db_info_df <- db_list %>% 
  map_dfr(get_size_and_rnacentral, .id = "mgdb")
db_info_df %>% mutate(frac_w_rnaids = (n_entries_rnaid)/n_entries)
```

- RNAcentral id overlap
```{r}
get_rnacentral_ids <- function(mgdb){
   if("rnacentral_ids" %in% taxa_columns(mgdb)){
    rna_filt <- mgDb_taxa(mgdb) %>% 
      filter(!is.na(rnacentral_ids)) 
  } else {
    rna_filt <- mgDb_taxa(mgdb) %>% 
      filter(!is.na(RNAcentralID)) 
  }
  rna_filt %>% 
    select(contains("central")) %>% 
    collect() %>% 
    distinct() %>% 
    .[[1,]]
}

## This is really slow ...
rna_central_id_list <- db_list %>% 
  map(get_rnacentral_ids)
```

```{r}
venn(rna_central_id_list)
```

# _Paenibacillus_ Sequence and Taxonomy Data
First subset the databases using the `mgDb_select()` function.
Then we will summarize the taxonomy data using functions from `tidyverse` package, specifically `dplyr`, `stringr` and `forcats` functions for manipulating `data.frames`, `strings`, and `factor` vectors respectively.

__TODO__ Clean-up do not want to load both taxa and seq into memory
```{r subsetDb, echo = TRUE}
paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_silva <- metagenomeFeatures::mgDb_select(slv128.1MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
paeni_16S_rdp <- metagenomeFeatures::mgDb_select(rdp11.5MgDb,
                          type = c("taxa"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
```

```{r getCounts, echo = FALSE}
## Per genus count data - greengeness
tidy_gg_taxa <- paeni_16S_greengenes %>%
    ## cleaning up species names
    mutate(Species = if_else(Species == "s__", "Unassigned", Species),
           Species = str_replace(Species, "s__",""))

taxa_df_greengenes <- tidy_gg_taxa %>%
    group_by(Species) %>%
    summarise(Count = n()) %>%
    ungroup() %>%
    mutate(Species = fct_reorder(Species, Count))

## Count info for text
total_otus_greengenes <- sum(taxa_df_greengenes$Count)
unassigned_idx_greengenes <- taxa_df_greengenes$Species == "Unassigned"
no_species_assignment_greengenes <- taxa_df_greengenes$Count[unassigned_idx_greengenes]

## Per genus count data - SILVA
tidy_slv_taxa <- paeni_16S_silva %>% rowwise() %>% 
  mutate(new = str_split(Species, boundary("word")),Species = paste(new[1],new[2])) %>% 
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

taxa_df_silva <- tidy_slv_taxa %>% 
  group_by(Species) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  mutate(Species = fct_reorder(Species, Count))

## Count info for text
total_otus_silva <- sum(taxa_df_silva$Count)
unassigned_idx_silva <- taxa_df_silva$Species == "Unassigned"
no_Species_assignment_silva <- taxa_df_silva$Count[unassigned_idx_silva]

## Per genus count data - RDP
tidy_rdp_taxa <- paeni_16S_rdp %>% rowwise() %>% 
  dplyr::rename(Species = species, Genus = genus) %>% 
  mutate(new = str_split(Species, boundary("word")), Species = paste(new[1],new[2])) %>% 
  mutate(new = NULL) %>%
  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
  mutate(Species = str_remove(Species, "Paenibacillus "))

taxa_df_rdp <- tidy_rdp_taxa %>% 
  group_by(Species) %>%
  summarise(Count = n()) %>%
  ungroup() %>%
  mutate(Species = fct_reorder(Species, Count))

## Count info for text
total_otus_rdp <- sum(taxa_df_rdp$Count)
unassigned_idx_rdp <- taxa_df_rdp$Species == "Unassigned"
no_Species_assignment_rdp <- taxa_df_rdp$Count[unassigned_idx_rdp]
#print (paste0("Number of sequences with unassigned species label: ", taxa_df_rdp$Count[taxa_df_rdp$Species == "Unassigned"]))
```

__TODO__ Revise text - all three databases
For the Greengenes 13.5 database, there are a total of `r total_otus_greengenes` sequences classified as `r nlevels(taxa_df_greengenes$Species)` species in the Genus _Paenibacillus_.
The number of sequences assigned to specific _Paenibacillus_ species range from 199 _Paenibacillus amylolyticus_ to 2 _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)).
Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment_greengenes`.

Similarly, we extracted _Paenibacillus_ sequences from Silva 128 database. Fig. \@ref(fig:speciesCountSilva) shows the number of species in the Genus _Paenibacillus_ in the Silva database.

For the RDP 11.5 database, all the sequences belonging to _Paenibacillus_ genus were either uncultured, unidentified, or unassigned.

```{r speciesCount, fig.cap = "Number of sequences in the Greengenes database assigned to species in the genus _Paenibacillus_ for the SILVA and Greengenes databases."}
combined_taxa_df <- bind_rows(list(Greengenes = taxa_df_greengenes, 
                                   SILVA = taxa_df_silva, 
                                   RDP = taxa_df_rdp), 
                              .id = "Database") %>% 
  group_by(Species) %>% 
  mutate(max_count = max(Count, na.rm = TRUE)) %>% 
  ungroup() %>% 
  top_n(60, max_count) %>% 
  mutate(Species = fct_reorder(Species, max_count))

combined_taxa_df %>% 
  group_by(Species) %>% 
  mutate(ymin = min(Count),
         ymax = max(Count)) %>% 
ggplot() +
    geom_linerange(aes(x = Species, ymin = ymin + 1, ymax = ymax + 1)) + 
    geom_point(aes(x = Species, y = Count + 1, color = Database)) + 
    labs(y = "Number of sequences") +
  scale_y_log10() + 
    coord_flip() +
    theme_bw() +
    theme(axis.text.y = element_text(face = "italic"),
          legend.position = "bottom") 
```

## Taxonomic resolution
Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
To differentiate between Species the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
For our taxonomic resolution analysis, we used pattern matching to extract the V12 and V4 regions of the 16S rRNA sequences.
We then generate a pairwise distance matrix for the two regions and compare the within and between Species pairwise distances.
Here, we work with sequences only in Greengenes 13.5 database for simplicity, the same analysis can be done with any other database.

For our _in-silico_ PCR we will use the following PCR primers:

|Region |Direction |Primer                             |
|:------|:-------- |:----------------------------------|
|V12    |Forward   | 27F - AGAGTTTGATCATGGCTCAG        |
|       |Reverse   | 336R - CACTGCTGCSYCCCGTAGGAGTCT   |
|V4     |Forward   | 515F - GTGCCAGCMGCCGCGGTAA        |
|       |Reverse   | 806R - GGACTACHVGGGTWTCTAAT       |

### Generating Distance Matrix for 16S Amplicons
Demonstrate process using gg and V12. 
Code hidden for other databases and V4, see Rmarkdown for full code, __PATH to Rmarkdown in repo__.

Extracting the V12 region from the database sequences, only sequences with containing both forward and reverse primers are included in the analysis.

```{r trimSeqs, echo = TRUE}
forward_primer <- "AGAGTTTGATCATGGCTCAG"
## reverse complementing reverse primer
reverse_primer <- DNAString("CACTGCTGCSYCCCGTAGGAGTCT") %>%
    reverseComplement() %>%
    as.character()

## Finding sequeces with forward primer
forward_match <- Biostrings::vmatchPattern(forward_primer,
                                           subject = paeni_16S_greengenes$seq,
                                           max.mismatch = 2) %>%
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## Finding sequences with reverse primer
reverse_match <- Biostrings::vmatchPattern(reverse_primer,
                                           subject = paeni_16S_greengenes$seq,
                                           max.mismatch = 2,
                                           fixed = FALSE) %>%
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## sequences with both forward and reverse primers
seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
seqs_to_use <- names(paeni_16S_greengenes$seq) %in% seqs_to_use_ids

## Trimming sequences with both primers
paeni_V12 <- TrimDNA(paeni_16S_greengenes$seq[seqs_to_use],
                     leftPatterns = forward_primer,
                     rightPatterns = reverse_primer,
                     type = "both")

## Excluding seqs with lenght 0
paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
```

Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.

```{r align, echo = TRUE}
v12_align <- AlignSeqs(paeni_V12[[2]], verbose = FALSE)
```

The resulting alignment can be viewed using the `BrowseSeqs` function in the `DECIPHER` package.

```{r eval = FALSE, echo = TRUE}
BrowseSeqs(v12_align)
```

Generating pairwise distance matrix using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.

```{r v12Anno, echo = TRUE}
v12_dist <- DistanceMatrix(v12_align,
                           correction = "none",
                           verbose = FALSE,
                           includeTerminalGaps = FALSE)
```

### Multiple Database Comparison
```{r distDFfuns, echo = FALSE, message = FALSE}
get_amp <- function(paeni_16S, forward_primer, reverse_primer){

    ## Finding sequeces with forward primer
    forward_match <- Biostrings::vmatchPattern(forward_primer, 
                                               subject = paeni_16S, 
                                               max.mismatch = 2) %>% 
        as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

    ## Finding sequences with reverse primer 
    reverse_match <- Biostrings::vmatchPattern(reverse_primer, 
                                               subject = paeni_16S, 
                                               max.mismatch = 2,
                                               fixed = FALSE) %>% 
        as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

    ## sequences with both forward and reverse primers 
    seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
    seqs_to_use <- names(paeni_16S) %in% seqs_to_use_ids

    ## Trimming sequences with both primers
    paeni_amp <- TrimDNA(paeni_16S[seqs_to_use],
                         leftPatterns = forward_primer,
                         rightPatterns = reverse_primer,
                         type = "both")

    ## Excluding seqs with lenght 0
    paeni_amp <- paeni_amp[[2]][width(paeni_amp[[2]]) != 0]

    ## Returning target seqs
    paeni_amp
}

get_dist <- function(amp_seqs){
  aligned_seqs <- AlignSeqs(amp_seqs, verbose = FALSE)
  amp_dist <- DistanceMatrix(aligned_seqs,
                            correction = "none",
                            verbose = FALSE,
                            includeTerminalGaps = FALSE)

  Key_ord <- data.frame(Keys = rownames(amp_dist), 
                        Keys_ord = 1:nrow(amp_dist))
  Key2_ord <- data.frame(Keys2 = rownames(amp_dist), 
                         Keys2_ord = 1:nrow(amp_dist))

  ## Creating a data frame for exploratory analysis
  amp_dist_df <- amp_dist %>%
      as.data.frame() %>%
      rownames_to_column(var = "Keys") %>%
      gather("Keys2","distance", -Keys) %>%
      left_join(Key_ord) %>% 
      left_join(Key2_ord) %>% 
      filter(Keys_ord < Keys2_ord) %>% 
      select(-Keys_ord, -Keys2_ord)
}

get_tidy_dist_df <- function(mgdb, tax_df, forward_primer, reverse_primer){
    ## Get seqs and taxa
    paeni_16S <- metagenomeFeatures::mgDb_select(mgdb,
                              type = "seq",
                              keys = tax_df$Keys,
                              keytype = "Keys")

    ## Subset seqs
    amp_seqs <- get_amp(paeni_16S, forward_primer, reverse_primer)

    ## Get distance DF 
    amp_dist_df <- get_dist(amp_seqs)

    ## get distance df
    amp_dist_anno_df <- amp_dist_df %>%
      left_join(tax_df) %>%
      left_join(tax_df,by = c("Keys2" = "Keys")) %>%
      dplyr::rename(Keys_Species = Species.x, 
                    Keys2_Species = Species.y) %>%
      mutate(group_comp = if_else(Keys_Species == Keys2_Species, 
                                  "within","between"))

    ## Return tidy dataframe
    amp_dist_anno_df
}
```

```{r getDistDF, echo = FALSE, message = FALSE}
### GG
gg_for_seq_comp <- tidy_gg_taxa %>% filter(!is.na(rnacentral_ids), Species != "Unassigned")

### SILVA
slv_for_seq_comp <- tidy_slv_taxa %>% filter(!is.na(RNAcentralID), Species != "Unassigned")

### RDP
rdp_for_seq_comp <- tidy_rdp_taxa %>% filter(!is.na(RNAcentralID), Species != "Unassigned")

## V12 Amplicon
forward_primer <- "AGAGTTTGATCATGGCTCAG"
## reverse complementing reverse primer
reverse_primer <- DNAString("CACTGCTGCSYCCCGTAGGAGTCT") %>% 
    reverseComplement() %>% 
    as.character()

taxa_list <- list(gg = gg_for_seq_comp, 
                  slv = slv_for_seq_comp, 
                  rdp = rdp_for_seq_comp)

multidb_dist_V12_df <- map2_dfr(db_list, taxa_list,
                                get_tidy_dist_df,
                                forward_primer,
                                reverse_primer,
                                .id = "Database")

## V4 Amplicon
forward_primer <- "GTGCCAGCMGCCGCGGTAA"
## reverse complementing reverse primer
reverse_primer <- "ATTAGAWACCCBDGTAGTCC"

multidb_dist_V4_df <- map2_dfr(db_list, taxa_list,
                                get_tidy_dist_df,
                                forward_primer,
                                reverse_primer,
                                .id = "Database")

## Combine into a single dataframe 
multidb_dist_df <- bind_rows(
  list(V12 = multidb_dist_V12_df, 
       V4  = multidb_dist_V4_df),
  .id = "amplicon")
```

__TODO__ Check about removal
```{r}
# ## Excluding outlier sequence "329842" - mean pairwise distance to all other
# ## sequences is 0.2
# v4_dist_anno_filt <- filter(v4_dist_anno_df, 
#                              Keys != "329842", 
#                              Keys2 != "329842")
```

#### Amplicon Sequence Lengths
The trimmed sequence length varies for forward and reverse primers resulting in varying amplicon sizes for both the V12 and V4 amplicons (Fig. \@ref(fig:trimInfo)).

```{r trimInfo, fig.cap = "Primer trimmed sequence, amplicon, length and start and end positions relative to full length sequences for the V12 and V4 regions."}
# list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>%
#     map_dfr(as.data.frame, .id = "amplicon") %>% 
#     filter(width != 0) %>% 
#     gather("key","value", -names, -amplicon) %>%
#     mutate(key = if_else(key == "width", "length", key),
#            key = factor(key, levels = c("length","start","end"))) %>%
#     mutate(position_type = if_else(key == "length",
#                                    "length", "position")) %>%
#     ggplot() +
#     geom_histogram(aes(x = value, fill = key), bins = 100) +
#     facet_grid(amplicon~position_type, scales = "free") +
#     theme_bw()
```

#### Genus Level Comparison

Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable forclassifying members of the _Paenibacillus_ genus to the Species level (Fig. \@ref(fig:pairDist)).
Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
It is important also to consider that the majority of sequences in the database were only classified to the genus level.
Species-level information for these sequences might yield results that are inconsistent with our analysis.
Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.

```{r pairDist, fig.cap = "Distribution of within and between Species pairwise distances for the V12 and V4 16S rRNA region. Sequences not classified to the Species level were excluded from the analysis."}
multidb_dist_df  %>% 
  # filter(Database == "gg") %>% 
    ggplot(aes(x = group_comp, y = distance, fill = Database)) +
    geom_boxplot() +
    facet_grid(.~amplicon) +
    labs(x = "Species Comparison", fill = "Database") +
        theme_bw() +
    theme(legend.position = "bottom")
```

#### Species level comparison
While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ Species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
In general though for Species levels classification the maximum within Species distance should be less than the minimum between Species distance.
For example as the maximum within Species pairwise distance for _P. lentimorbus_ is 0.13 and the minimum between Species pairwise distance for _P. lentimorbus_ and _P. alvei_ is 0.08 (Fig. \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.

__REVISE figure legend__
```{r distHeatMap, fig.cap = "Pairwise distances between _Paenibacillus_ Species (A) V12 and (B) V4 amplicon regions. Fill color indicates the mean pairwise sequence distance within and between Species. The text indicates the maximum pairwise distance for within-Species comparisons, values along the diagonal, and maximum pairwise distance for between Species comparisons. Different number of Species are included in the V12 and V4 plots as there are no full-length \\textit{P. chondroitinus} sequences with the V12 primer in the database.", fig.height = 8}
multidb_dist_df %>% 
  filter(Keys_Species %in% c("polymyxa", "amylolyticus", "lautus","chondroitinus"),
         Keys2_Species %in% c("polymyxa", "amylolyticus", "lautus","chondroitinus")) %>% 
    ggplot() +
    geom_boxplot(aes(x = amplicon,
                    y = distance,
                    fill = Database)) +
    facet_grid(Keys_Species~Keys2_Species) +
   theme_bw()
```

## Conclusion
Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`class object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).

********************************************************************************

## Session Information
### System Information
```{r warning = FALSE}
s_info <- devtools::session_info(include_base = FALSE)
pander::pander(s_info$platform)
```

### Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>%
    knitr::kable(booktabs = TRUE)
```

<!-- __Potential reference__ https://www.biorxiv.org/content/early/2018/03/25/288654?%3Fcollection= -->
<!-- Sequencing of the 16S ribosomal RNA (rRNA) gene and the fungal Internal Transcribed Spacer (ITS) region is widely used to survey microbial communities. Specialized ribosomal sequence databases have been developed to support this approach including Greengenes, SILVA and RDP. Most taxonomy annotations in these databases are predictions from sequence rather than authoritative assignments based on studies of type strains or isolates. Here, I investigate the error rates of taxonomy annotations in these databases. I found 253,485 sequences with conflicting annotations in SILVA v128 and Greengenes v13.5 at ranks up to phylum (9,644 conflicts), indicating that the annotation error rate in these databases is ~15%. I found that 34% of non-singleton genera have overlapping subtrees in the Greengenes tree from 2001 according to the RDP taxonomy, most of which are probably due to branching order errors in the Greengenes tree, which is therefore an unreliable guide to phylogeny. Using a blinded test, I estimated that the annotation error rate of the RDP database is ~10%. -->

### 2018-11-08 inst/manuscript/supplemental_material.Rmd
revised text

@@
 title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
-subtitle: "_Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions."
 author: "Nathan D. Olson and Nidhi Shah"
@@
 
+- Multiple 16S rRNA databases
+- Can perform analysis on an individual database
+- with metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
 Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
@@
 
+While the databases provide a consistent interface to the different 16S rRNA databases, the databases use different approaches to formating the taxonomic names.
+As a result the majority of the code used to produce this document is tidying the data.
+This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository, __ADD URL__
+
 # Required Packages
@@
 library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
-# library(silva128.1MgDb, lib.loc = "~/Documents/Fall2018/installed_packages/"); packageVersion("silva128.1MgDb")
 library(silva128.1MgDb); packageVersion("silva128.1MgDb")
@@
 
+# 16S rRNA Database Comparison
 # Database Comparison
@@
 
-# _Paenibacillus_ Sequence and Taxonomy Data
-First subset the databases using the `mgDb_select()` function.
-Then we will summarize the taxonomy data using functions from `tidyverse` package, specifically `dplyr`, `stringr` and `forcats` functions for manipulating `data.frames`, `strings`, and `factor` vectors respectively.
+# _Paenibacillus_ Taxonomic Resolution
+For the taxonomic resolution analysis we will;
+1. characterize the taxonomic composition of the _Paenibacillus_ genus in the three databases,
+2. calculate pairwise distances between sequences for 30 species with the most sequences,
+3. evaluate pairwise distances within and between species.
 
-__TODO__ Clean-up do not want to load both taxa and seq into memory
+## Taxonomic Characterization
+First obtain the taxonomic information for _Paenibacillus_ genus from the three databases using the `mgDb_select()` function.
+
 ```{r subsetDb, echo = TRUE}
@@
 ```{r getCounts, echo = FALSE}
-## Per genus count data - greengeness
+## Per genus count data - greengenes ###########################################
 tidy_gg_taxa <- paeni_16S_greengenes %>%
@@
 
-## Per genus count data - SILVA
+## Per genus count data - SILVA ################################################
 tidy_slv_taxa <- paeni_16S_silva %>% rowwise() %>%
@@
 
-## Per genus count data - RDP
+## Per genus count data - RDP ##################################################
 tidy_rdp_taxa <- paeni_16S_rdp %>% rowwise() %>%
@@
 
-__TODO__ Revise text - all three databases
-For the Greengenes 13.5 database, there are a total of `r total_otus_greengenes` sequences classified as `r nlevels(taxa_df_greengenes$Species)` species in the Genus _Paenibacillus_.
-The number of sequences assigned to specific _Paenibacillus_ species range from 199 _Paenibacillus amylolyticus_ to 2 _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)).
-Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment_greengenes`.
-
-Similarly, we extracted _Paenibacillus_ sequences from Silva 128 database. Fig. \@ref(fig:speciesCountSilva) shows the number of species in the Genus _Paenibacillus_ in the Silva database.
-
-For the RDP 11.5 database, all the sequences belonging to _Paenibacillus_ genus were either uncultured, unidentified, or unassigned.
+- Relatively small overlap between the databases, multiple database analysis provide additional power.
+- The databases vary in the number of sequences for each species with some databases not having any sequences for some species.
+- Number of sequences and species for each database
+- Greengenes benefit - cleaner species names
+- SILVA benefit - additional sequences
+- RDP benefit - additional sequences
+- Issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_species_assignment_silva`, and RDP - `r no_species_assignment_rdp` __FIG-REF__.
 
@@
   ungroup() %>%
-  top_n(60, max_count) %>% 
+  top_n(30, max_count) %>%
   mutate(Species = fct_reorder(Species, max_count))
@@
 
-## Taxonomic resolution
+## Taxonomic Resolution
 Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
@@
 Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
-For our taxonomic resolution analysis, we used pattern matching to extract the V12 and V4 regions of the 16S rRNA sequences.
-We then generate a pairwise distance matrix for the two regions and compare the within and between Species pairwise distances.
-Here, we work with sequences only in Greengenes 13.5 database for simplicity, the same analysis can be done with any other database.
 
-For our _in-silico_ PCR we will use the following PCR primers:
+### Generating Distance Matrix for 16S Amplicons
+Here, we will demonstrate the process for extracting the target amplicon region and calculating pairwise distances using the Greengenes 13.5 database for simplicity.
+See the source Rmarkdown file for code used calculate pairwise distances for other databases and the V4 region as well as tidying the data for downstream analysis.
+
+The V12 and V4 regions were extracted from the database sequences using pattern matching.
+Some sequences do not contain both forward and reverse primers as not all the  16S rRNA sequences are not full length.
+Only sequences with both forward and reverse primers are included in the analysis.
+
+The following PCR primers were used for our _in-silico_ PCR:
 
@@
 
-### Generating Distance Matrix for 16S Amplicons
-Demonstrate process using gg and V12. 
-Code hidden for other databases and V4, see Rmarkdown for full code, __PATH to Rmarkdown in repo__.
-
-Extracting the V12 region from the database sequences, only sequences with containing both forward and reverse primers are included in the analysis.
-
 ```{r trimSeqs, echo = TRUE}
@@
     as.character()
+
+## Selecting Paenibacillus taxa and seq
+paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
+                          type = c("taxa", "seq"),
+                          keys = "Paenibacillus",
+                          keytype = "Genus")
 
@@
 
-## Excluding seqs with lenght 0
+## Excluding seqs with length 0
 paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
@@
 
-Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.
+For more accurate distance estimates a multiple sequence alignment is used to calculate pairwise distances.
+We will use the `AlignSeqs` function in the `DECIPHER` package to generate the multiple sequence alignment.
 
@@
 
-Generating pairwise distance matrix using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.
+Next a pairwise distance matrix is generated using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.
 
@@
 ### Multiple Database Comparison
+Now that we have our pairwise distance matrix we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species.
+We will first compare the distribution of within and between species pairwise distances for the genus.
+Then look at the pairwise distances within species __XYZ__ compared distances between __XYZ__ and other _Paenibacillu_ species.
+To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the __###__ species with the most sequences.
+With more sequences we will have better estimates for within species pairwise distances but is likely to result in the exclusion of closely related species.
+
 ```{r distDFfuns, echo = FALSE, message = FALSE}
@@
 
-#### Amplicon Sequence Lengths
-The trimmed sequence length varies for forward and reverse primers resulting in varying amplicon sizes for both the V12 and V4 amplicons (Fig. \@ref(fig:trimInfo)).
-
-```{r trimInfo, fig.cap = "Primer trimmed sequence, amplicon, length and start and end positions relative to full length sequences for the V12 and V4 regions."}
-# list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>%
-#     map_dfr(as.data.frame, .id = "amplicon") %>% 
-#     filter(width != 0) %>% 
-#     gather("key","value", -names, -amplicon) %>%
-#     mutate(key = if_else(key == "width", "length", key),
-#            key = factor(key, levels = c("length","start","end"))) %>%
-#     mutate(position_type = if_else(key == "length",
-#                                    "length", "position")) %>%
-#     ggplot() +
-#     geom_histogram(aes(x = value, fill = key), bins = 100) +
-#     facet_grid(amplicon~position_type, scales = "free") +
-#     theme_bw()
-```
-
 #### Genus Level Comparison
 
-Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable forclassifying members of the _Paenibacillus_ genus to the Species level (Fig.
+Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable for classifying members of the _Paenibacillus_ genus to the Species level (Fig.
 \@ref(fig:pairDist)).
+__TO ADD - Conclusions from multiple database analysis__.
 Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
-It is important also to consider that the majority of sequences in the database were only classified to the genus level.
-Species-level information for these sequences might yield results that are inconsistent with our analysis.
+It is important also to consider that we only included sequences with species assignments and the species with the most sequences in the database.
+Species-level information for the unclassified sequences and inclusion of less well represented species might yield results that are inconsistent with our analysis.
 Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.
@@
 #### Species level comparison
+Add text for new figure
+
+<!--
 While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
@@
 \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.
+-->
 
@@
 
-## Conclusion
-Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
-The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`class object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
+# Conclusion
+__TO ADD - Statement about database overlap and using RNAcentralIDs__
+Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, _silva128.1MgDb_, _ribosomaldatabaseproject11.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
+The approach used here can easily be extended to use evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
 
@@
 
-## Session Information
-### System Information
+# Session Information
+## System Information
 ```{r warning = FALSE}
@@
 
-### Package Versions
+## Package Versions
 ```{r}

### 2018-11-10 inst/manuscript/manuscript_supplemental-initial-submission.Rmd
renamed files

---
title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
author: "Nate Olson"
date: "`r Sys.Date()`"
output: bookdown::pdf_document2
---
<!--
__Potential reference__ https://www.biorxiv.org/content/early/2018/03/25/288654?%3Fcollection=
Sequencing of the 16S ribosomal RNA (rRNA) gene and the fungal Internal Transcribed Spacer (ITS) region is widely used to survey microbial communities. Specialized ribosomal sequence databases have been developed to support this approach including Greengenes, SILVA and RDP. Most taxonomy annotations in these databases are predictions from sequence rather than authoritative assignments based on studies of type strains or isolates. Here, I investigate the error rates of taxonomy annotations in these databases. I found 253,485 sequences with conflicting annotations in SILVA v128 and Greengenes v13.5 at ranks up to phylum (9,644 conflicts), indicating that the annotation error rate in these databases is ~15%. I found that 34% of non-singleton genera have overlapping subtrees in the Greengenes tree from 2001 according to the RDP taxonomy, most of which are probably due to branching order errors in the Greengenes tree, which is therefore an unreliable guide to phylogeny. Using a blinded test, I estimated that the annotation error rate of the RDP database is ~10%.
 -->
# _Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions.

### Background
16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance analysis.
A limitation to 16S rRNA amplicon sequencing is a lack of taxonomic resolution, where organisms are only identifiable to the genus or family level.
We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus. 
While similar to determining whether a sequence represents a novel species, here we are only interested in determining whether the 16S rRNA region of interest contains sufficient information for species-level taxonomic assignment.
Taxonomic resolution varies by clade and amplicon regions. 
Though the extent to which taxonomic resolition varies is not well characterized.

Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions. Originally classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s. 
_Paenibacillus_ spp. are facultative anaerobic bacteria present in a variety of environments including the soil, water, and can act as opportunistic pathogens in humans [@ouyang2008]. 
_Paenibacillus_ spp. will play an important role in sustainable agricultural industries [@grady2016]. As such, appropriate speciation is of interest. 
The V12 and V4 region were used as they represent two commonly used amplicons for 16S rRNA marker-gene surveys. 
We will use the Greengenes 13.5 database, accessed using the _greengenes13.5MgDb_ annotation package for our analysis of the _Paenibacillus_ genus. 
The Greengenes 13.5 database is used for demonstration purposes but the other `MgDb` annotation packages can also be used; RDP 11.5 - `ribosomaldatabaseproject11.5MgDb or SILVA 128.1 - `silva128.1MgDb`.

### Required Packages 
In addition to _metagenomeFeatures_ and _greengenes13.5MgDb_ the _DECIPHER_, _tidyverse_, and _ggpubr_ packages are also used in the following analysis. 
Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@Wright2016-mo].
The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse;@ggpubr].

```{r message = FALSE, echo = TRUE}
library(tidyverse); packageVersion("tidyverse")
library(ggpubr); packageVersion("ggpubr")
library(DECIPHER); packageVersion("DECIPHER")
library(metagenomeFeatures); packageVersion("metagenomeFeatures")
library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
```

### _Paenibacillus_ Sequence and Taxonomy Data
We first subset the Greeengenes 13.5 database using the `mgDb_select` function.
Then summarize the taxonomy data using functions from `tidyverse` package, specifically `dplyr`, `stringr` and `forcats` functions for manipulating `data.frames`, `strings`, and `factor` vectors respectively.
<!-- Loading metagenomeFeatues and greengenes db -->

```{r subsetDb, echo = TRUE}
paeni_16S <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
                          type = c("taxa","seq"),
                          keys = "Paenibacillus",
                          keytype = "Genus")
```

```{r getCount, echo = TRUE}
## Per genus count data
taxa_df <- paeni_16S$taxa %>%
    ## cleaning up species names
    mutate(Species = if_else(Species == "s__", "Unassigned", Species),
           Species = str_replace(Species, "s__","")) %>%
    group_by(Species) %>%
    summarise(Count = n()) %>%
    ungroup() %>%
    mutate(Species = fct_reorder(Species, Count))

## Count info for text
total_otus <- sum(taxa_df$Count)
unassigned_idx <- taxa_df$Species == "Unassigned"
no_species_assignment <- taxa_df$Count[unassigned_idx]
```

For the Greengenes 13.5 database, there are a total of `r total_otus` sequences classified as `r nlevels(taxa_df$Species)` species in the Genus _Paenibacillus_.
The number of sequences assigned to specific _Paenibacillus_ species range from 199 _Paenibacillus amylolyticus_ to 2 _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)).
Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment`.

```{r speciesCount, fig.cap = "Number of sequences assigned to species in the genus _Paenibacillus_."}
taxa_df %>%
ggplot() +
    geom_bar(aes(x = Species, y = Count), stat = "identity") +
    geom_text(aes(x = Species, y = Count, label = Count), 
              nudge_y = 75) +
    labs(y = "Number of OTUs") +
    coord_flip() +
    theme_bw() +
    theme(axis.text.y = element_text(face = "italic"))
```

### Taxonomic resolution
Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ species by comparing within and between species amplicon pairwise distance for the V12 and V4 regions.
To differentiate between species the pairwise distances for within-species amplicon sequences must be less than the between species distances.
Additionally ,the difference in amplicon sequence pairwise distances between and within species must be greater than the sequencing error rate to detect the difference.
For our taxonomic resolution analysis, we used pattern matching to extract the V12 and V4 regions of the 16S rRNA sequences.
We then generate a pairwise distance matrix for the two regions and compare the within and between species pairwise distances.

For our _in-silico_ PCR we will use the following PCR primers:

|Region |Direction |Primer                             |
|:------|:-------- |:----------------------------------|
|V12    |Forward   | 27F - AGAGTTTGATCATGGCTCAG        |
|       |Reverse   | 336R - CACTGCTGCSYCCCGTAGGAGTCT   |
|V4     |Forward   | 515F - GTGCCAGCMGCCGCGGTAA        |
|       |Reverse   | 806R - GGACTACHVGGGTWTCTAAT       |

#### V12
Extracting the V12 region from the database sequences, only sequences with containing both forward and reverse primers are included in the analysis.

```{r trimSeqs, echo = TRUE}
forward_primer <- "AGAGTTTGATCATGGCTCAG"
## reverse complementing reverse primer
reverse_primer <- DNAString("CACTGCTGCSYCCCGTAGGAGTCT") %>% 
    reverseComplement() %>% 
    as.character()

## Finding sequeces with forward primer
forward_match <- Biostrings::vmatchPattern(forward_primer, 
                                           subject = paeni_16S$seq, 
                                           max.mismatch = 2) %>% 
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## Finding sequences with reverse primer 
reverse_match <- Biostrings::vmatchPattern(reverse_primer, 
                                           subject = paeni_16S$seq, 
                                           max.mismatch = 2,
                                           fixed = FALSE) %>% 
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## sequences with both forward and reverse primers 
seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
seqs_to_use <- names(paeni_16S$seq) %in% seqs_to_use_ids

## Trimming sequences with both primers
paeni_V12 <- TrimDNA(paeni_16S$seq[seqs_to_use],
                     leftPatterns = forward_primer,
                     rightPatterns = reverse_primer,
                     type = "both")

## Excluding seqs with lenght 0
paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
```

Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.

```{r align, echo = TRUE}
v12_align <- AlignSeqs(paeni_V12[[2]], verbose = FALSE)
```

The resulting alignment can be viewed using the `BrowseSeqs` function in the `DECIPHER` package.

```{r eval = FALSE, echo = TRUE}
BrowseSeqs(v12_align)
```

Generating pairwise distance matrix using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.

```{r v12Anno, echo = TRUE}
v12_dist <- DistanceMatrix(v12_align,
                           correction = "none",
                           verbose = FALSE,
                           includeTerminalGaps = FALSE)

v12_dist_df <- v12_dist %>%
    as.data.frame() %>%
    rownames_to_column(var = "Keys") %>%
    gather("Keys2","distance", -Keys) %>%
    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>%
    filter(Keys < Keys2) %>%
    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))

tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species")
v12_dist_anno_df <- v12_dist_df %>%
    left_join(tax_df) %>%
    left_join(tax_df,by = c("Keys2" = "Keys")) %>%
    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>%
        mutate(group_comp = if_else(Keys_Species == Keys2_Species, 
                                    "within","between")) %>%
    filter(Keys_Species != "s__", Keys2_Species != "s__")
```

#### V4
For the V4 region, we will use the same approach, extract amplicon region, filter extracted sequences based on amplicon length, generate pairwise distance matrix using a multiple sequence alignment, and then evaluate pairwise distances.

```{r v4Anno, echo = TRUE}
## Finding sequeces with forward primer
forward_match <- Biostrings::vmatchPattern("GTGCCAGCMGCCGCGGTAA", 
                                           subject = paeni_16S$seq,
                                           fixed = FALSE) %>% 
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## Finding sequences with reverse primer 
reverse_match <- Biostrings::vmatchPattern("ATTAGAWACCCBDGTAGTCC", 
                                           subject = paeni_16S$seq, 
                                           fixed = FALSE) %>% 
    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")

## sequences with both forward and reverse primers 
seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
seqs_to_use <- names(paeni_16S$seq) %in% seqs_to_use_ids

## Extract amplicon region
paeni_V4 <- TrimDNA(paeni_16S$seq[seqs_to_use],
                               leftPatterns = "GTGCCAGCMGCCGCGGTAA",
                               rightPatterns = "ATTAGAWACCCBDGTAGTCC",
                               type = "both")

## Excluding seqs with lenght 0
paeni_V4_seqs <- paeni_V4[[2]][width(paeni_V4[[2]]) != 0]

### Calculate distance matrix from multiple sequence alignment
v4_align <- AlignSeqs(paeni_V4_seqs, verbose = FALSE)
v4_dist <- DistanceMatrix(v4_align,
                          correction = "none",
                          verbose = FALSE,
                          includeTerminalGaps = FALSE)

## Creating a data frame for exploratory analysis
v4_dist_df <- v4_dist %>%
    as.data.frame() %>%
    rownames_to_column(var = "Keys") %>%
    gather("Keys2","distance", -Keys) %>%
    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>%
    filter(Keys < Keys2) %>%
    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))

tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species")
v4_dist_anno_df <- v4_dist_df %>%
    left_join(tax_df) %>%
    left_join(tax_df,by = c("Keys2" = "Keys")) %>%
    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>%
    mutate(group_comp = if_else(Keys_Species == Keys2_Species, 
                                "within","between")) %>%
    filter(Keys_Species != "s__", Keys2_Species != "s__")

## Excluding outlier sequence "329842" - mean pairwise distance to all other
## sequences is 0.2
 v4_dist_anno_filt <- filter(v4_dist_anno_df, 
                             Keys != "329842", 
                             Keys2 != "329842")
```

#### Amplicon Sequence Lengths
The trimmed sequence length varies for forward and reverse primers resulting in varying amplicon sizes for both the V12 and V4 amplicons (Fig. \@ref(fig:trimInfo)).

```{r trimInfo, fig.cap = "Primer trimmed sequence, amplicon, length and start and end positions relative to full length sequences for the V12 and V4 regions."}
list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>%
    map_dfr(as.data.frame, .id = "amplicon") %>% 
    filter(width != 0) %>% 
    gather("key","value", -names, -amplicon) %>%
    mutate(key = if_else(key == "width", "length", key),
           key = factor(key, levels = c("length","start","end"))) %>%
    mutate(position_type = if_else(key == "length",
                                   "length", "position")) %>%
    ggplot() +
    geom_histogram(aes(x = value, fill = key), bins = 100) +
    facet_grid(amplicon~position_type, scales = "free") +
    theme_bw()
```

##### Genus Level Comparison

Pairwise distance is significantly different for within and between species comparisons indicating that the V12 and V4 regions are potentially suitable forclassifying members of the _Paenibacillus_ genus to the species level (Fig. \@ref(fig:pairDist)). 
Overall the V12 region had greater pairwise distances than V4 for both within and between species.
It is important also to consider that the majority of sequences in the database were only classified to the genus level. 
Species-level information for these sequences might yield results that are inconsistent with our analysis.
Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ species.

```{r pairDist, fig.cap = "Distribution of within and between species pairwise distances for the V4 16S rRNA region. Sequences not classified to the species level were excluded from the analysis."}
list("V4" = v4_dist_anno_filt, "V12" = v12_dist_anno_df) %>%
    bind_rows(.id = "amplicon")  %>%
    ggplot(aes(x = group_comp, y = distance, fill = group_comp)) +
    geom_boxplot(aes(x = group_comp, y = distance, fill = group_comp)) +
    facet_wrap(~amplicon) +
    labs(x = "Amplicon", fill = "Species Comparison") +
        theme_bw() +
    theme(legend.position = "bottom")
```

##### Species level comparison
While the overall pairwise distance is greater between species than within species for the _Paenibacillus_ genus, it is important to understand how the within and between species pairwise distances compare for individual species.
The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
In general though for species levels classification the maximum within species distance should be less than the minimum between species distance.
For example as the maximum within species pairwise distance for _P. lentimorbus_ is 0.13 and the minimum between species pairwise distance for _P. lentimorbus_ and _P. alvei_ is 0.08 (Fig. \@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two species is not possible.

```{r distHeatMap, fig.cap = "Pairwise distances between _Paenibacillus_ species (A) V12 and (B) V4 amplicon regions. Fill color indicates the mean pairwise sequence distance within and between species. The text indicates the maximum pairwise distance for within-species comparisons, values along the diagonal, and maximum pairwise distance for between species comparisons. Different number of species are included in the V12 and V4 plots as there are no full-length \\textit{P. chondroitinus} sequences with the V12 primer in the database.", fig.height = 8}
v12_heat <- v12_dist_anno_df %>%
    group_by(Keys_Species, Keys2_Species, group_comp) %>%
    summarise(mean_dist = mean(distance, na.rm = TRUE),
              max_dist = max(distance),
              min_dist = min(distance)) %>%
    mutate(lab_dist = if_else(group_comp == "within",
                              max_dist, min_dist)) %>%
    ungroup() %>% 
    mutate(Keys_Species = str_remove(Keys_Species, "s__")) %>% 
    mutate(Keys2_Species = str_remove(Keys2_Species, "s__")) %>% 
    ggplot() +
    geom_raster(aes(x = Keys_Species,
                    y = Keys2_Species,
                    fill = mean_dist)) +
    geom_text(aes(x = Keys_Species, y = Keys2_Species, 
                  label = round(lab_dist,2)), size = 2, color = "white") +
    labs(x = "Species",y = "Species", fill = "Mean Dist") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))

v4_heat <- v4_dist_anno_df %>%
    filter(Keys != "329842", Keys2 != "329842") %>%
    group_by(Keys_Species, Keys2_Species, group_comp) %>%
    summarise(mean_dist = mean(distance),
              max_dist = max(distance),
              min_dist = min(distance)) %>%
    mutate(lab_dist = if_else(group_comp == "within", 
                              max_dist, min_dist)) %>%
    ungroup() %>% 
    mutate(Keys_Species = str_remove(Keys_Species, "s__")) %>% 
    mutate(Keys2_Species = str_remove(Keys2_Species, "s__")) %>% 
    ggplot() +
    geom_raster(aes(x = Keys_Species,
                    y = Keys2_Species,
                    fill = mean_dist)) +
    geom_text(aes(x = Keys_Species, y = Keys2_Species, 
                  label = round(lab_dist,2)), size = 2, color = "white") +
    labs(x = "Species",y = "Species", fill = "Mean Dist") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))

ggpubr::ggarrange(v12_heat, v4_heat,labels = "AUTO", 
                  ncol = 1, nrow = 2,align = "v")
```

### Conclusion
Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, and other R packages, can be used to evaluate whether species-level taxonomic classification is possible for a specific amplicon region.
The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`class object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).

********************************************************************************

## Session Information
### System Information
```{r}
s_info <- devtools::session_info(include_base = FALSE)
pander::pander(s_info$platform)
```

### Package Versions
```{r}
s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>%
    knitr::kable(booktabs = TRUE)
```

### 2018-11-10 inst/manuscript/manuscript_supplemental.Rmd
renamed files

--- 2018-03-29 inst/manuscript/manuscript_supplemental.Rmd
@@
 ---
-title: "_Paenibacillus_ species resolution for 16S rRNA V12 and V4 regions."
-author: "Nate Olson"
+title: "Supplemental Material - _metagenomeFeatures_: An R package for working with 16S rRNA reference databases and marker-gene survey feature data."
+author: "Nathan D. Olson and Nidhi Shah"
 date: "`r Sys.Date()`"
@@
 
-## Background
+# Background
 16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance analysis.
@@
 We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus.
-Taxonomic resolution can vary by clade and amplicon regions, though the extent to which it varies is not well characterized.
+We are interested in determining whether the 16S rRNA region of interest contains sufficient information for species-level taxonomic assignment.
+Taxonomic resolution varies by clade and amplicon regions.
+However, the extent to which taxonomic resolution varies is not well characterized.
 
-Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions
-Using the Greengenes 13.5, `greengenes13.5MgDb` we will first subset the database to obtain _Paenibacillus_ taxonomic and sequence data. 
-Next we will explore the _Paenibacillus_ species composition, then characterize the within and between species sequence similarity for the 16S rRNA gene V12 and V4 regions.
+There are multiple 16S rRNA databases whoes maintainers use different curation approaches resulting in different taxonomic and sequence compositions as well as being differently formatted. 
+Taxonomic or other 16S rRNA sequence analysis can perform analysis on an individual databases.
+With metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
+Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
+Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s.
+_Paenibacillus_ spp. are facultative anaerobic bacteria present in a variety of environments including the soil, water, and can act as opportunistic pathogens in humans [@ouyang2008].
+It has been shown that _Paenibacillus_ spp. will play an important role in sustainable agricultural industries [@grady2016].
+Thus, an appropriate speciation of this genus is of an interest to the community.
 
-## Required Packages  
-Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix. 
-The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures.
+We used V12 and V4 region as they represent two commonly used amplicons for 16S rRNA marker-gene surveys.
+We will use the Greengenes 13.5 database, accessed using the _greengenes13.5MgDb_, Silva 128 database, accessed using the _silva128.1MgDb_, and RDP 11.5 database, accessed using the _ribosomaldatabaseproject11.5MgDb_, as annotation packages for our analysis of the _Paenibacillus_ genus.
+First, we show the distribution of sequences belonging to _Paenibacillus_ in these three database, and then a way to combine different subsets of database sequences to build a custom database object.
+This helps investigators to pool sequences from different databases, to create an accurate and most representative database for their sample.
+This is especially helpful when a certain clade of interest is not very well represented in the databases.
+We then evaluate 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ species for V12 and V4 regions.
+In many analyses, the Greengenes 13.5 database is used for demonstration purposes but the other `MgDb` annotation packages such as RDP 11.5, SILVA 128, can also be used.
+
+While the databases provide a consistent interface to the different 16S rRNA databases, the databases use different approaches to formating the taxonomic names.
+As a result the majority of the code used to produce this document is tidying the data.
+This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository, __ADD URL__
+
+# Required Packages
+In addition to _metagenomeFeatures_, _greengenes13.5MgDb_, _ribosomaldatabaseproject11.5MgDb_, and _silva128.1MgDb_, the _DECIPHER_, _tidyverse_, and _ggpubr_ packages are also used in the following analysis.
+Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@Wright2016-mo].
+The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse;@ggpubr]. 
+See __Session Information Section__ for package version information.
 
 ```{r message = FALSE, echo = TRUE}
-library(tidyverse); packageVersion("tidyverse")
-library(ggpubr); packageVersion("ggpubr")
-library(DECIPHER); packageVersion("DECIPHER")
-library(metagenomeFeatures); packageVersion("metagenomeFeatures")
-library(greengenes13.5MgDb); packageVersion("greengenes13.5MgDb")
+library(tidyverse)
+library(ggpubr)
+library(gplots)
+library(DECIPHER)
+library(metagenomeFeatures)
+library(greengenes13.5MgDb)
+library(silva128.1MgDb)
+library(ribosomaldatabaseproject11.5MgDb)
 ```
 
-## _Paenibacillus_ Sequence and Taxonomy Data
-We first subset the Greeengenes 13.5 database using the `mgDb_select` function. 
-Then summarize the taxonomy data using the `tidyverse` package. 
-<!-- Loading metagenomeFeatues and greengenes db -->
+# 16S rRNA Database Comparison
+We have developed 16S rRNA MgDb annotation packages for three 16S rRNA databases, Greengenes, SILVA, and RDP. 
+RNAcentral a meta-database for non-coding RNA sequences provides cross database unique identifiers for sequences in the databases. 
+The number of sequences in the databases and fraction of sequences with RNAcentralIDs varies for the three databases (__TABLE REF__). 
+Using the RNAcentral IDs we can evaluate the overlap between the three databases (__FIG VENN__).
 
-```{r}
-paeni_16S <- mgDb_select(gg13.5MgDb, 
-                          type = c("taxa","seq"), 
+```{r echo = FALSE}
+get_rnacentral_entries <- function(mgdb){
+   if("rnacentral_ids" %in% taxa_columns(mgdb)){
+    rna_filt <- mgDb_taxa(mgdb) %>%
+      filter(!is.na(rnacentral_ids))
+  } else {
+    rna_filt <- mgDb_taxa(mgdb) %>%
+      filter(!is.na(RNAcentralID))
+  }
+  rna_filt
+}
+
+get_size_and_rnacentral <- function(mgdb){
+  db_size <- mgDb_taxa(mgdb) %>%
+    tally() %>% collect() %>% as.numeric()
+
+  db_wrna_id <- get_rnacentral_entries(mgdb) %>%
+    tally() %>% collect() %>% as.numeric()
+
+  ## return size and size with rna central ids
+  data_frame(Size = db_size,
+             n_entries_rnaid = db_wrna_id)
+}
+
+db_list <- list(Greengenes = gg13.5MgDb,
+                SILVA = slv128.1MgDb,
+                RDP = rdp11.5MgDb)
+
+db_info_df <- db_list %>%
+  map_dfr(get_size_and_rnacentral, .id = "Database")
+db_info_df %>% 
+  mutate(`Frac Ids` = (n_entries_rnaid)/Size) %>% 
+  dplyr::select(-n_entries_rnaid) %>% 
+  knitr::kable(digits = 2, format.args = list(big.mark = ","),
+               caption = "Number of sequences (size) and fraction of sequences with RNAcentral ids (frac ids) for the three 16S rRNA databases with MgDb annotation packages."
+               )
+```
+
+```{r echo = FALSE, warning = FALSE}
+get_rnacentral_ids <- function(mgdb){
+   if("rnacentral_ids" %in% taxa_columns(mgdb)){
+    rna_filt <- mgDb_taxa(mgdb) %>%
+      filter(!is.na(rnacentral_ids))
+  } else {
+    rna_filt <- mgDb_taxa(mgdb) %>%
+      filter(!is.na(RNAcentralID))
+  }
+  rna_filt %>%
+    select(contains("central")) %>%
+    collect() %>%
+    distinct() %>%
+    .[[1,]]
+}
+
+## This is really slow ...
+rna_central_id_list <- db_list %>%
+  map(get_rnacentral_ids)
+```
+
+```{r dbVenn, echo = FALSE, fig.cap = "Overlap in sequence composition between the three 16S rRNA databases with  MgDb annotation packages for sequences with RNAcentral IDs."}
+venn(rna_central_id_list,)
+```
+
+# _Paenibacillus_ Taxonomic Resolution
+For the taxonomic resolution analysis we will;
+1. characterize the taxonomic composition of the _Paenibacillus_ genus in the three databases,
+2. calculate pairwise distances between sequences for 30 species with the most sequences,
+3. evaluate pairwise distances within and between species.
+
+## Taxonomic Characterization
+First obtain the taxonomic information for _Paenibacillus_ genus from the three databases using the `mgDb_select()` function.
+
+```{r subsetDb, echo = TRUE}
+paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
+                          type = c("taxa"),
+                          keys = "Paenibacillus",
+                          keytype = "Genus")
+paeni_16S_silva <- metagenomeFeatures::mgDb_select(slv128.1MgDb,
+                          type = c("taxa"),
+                          keys = "Paenibacillus",
+                          keytype = "Genus")
+paeni_16S_rdp <- metagenomeFeatures::mgDb_select(rdp11.5MgDb,
+                          type = c("taxa"),
                           keys = "Paenibacillus",
@@
 
-```{r}
-## Per genus count data
-taxa_df <- paeni_16S$taxa %>% 
+```{r getCounts, echo = FALSE, warning = FALSE, message = FALSE}
+## Per genus count data - greengenes ###########################################
+tidy_gg_taxa <- paeni_16S_greengenes %>%
     ## cleaning up species names
     mutate(Species = if_else(Species == "s__", "Unassigned", Species),
-           Species = str_replace(Species, "s__","")) %>% 
+           Species = str_replace(Species, "s__",""))
+
+taxa_df_greengenes <- tidy_gg_taxa %>%
     group_by(Species) %>%
@@
 ## Count info for text
-total_otus <- sum(taxa_df$Count)
-no_species_assignment <- taxa_df$Count[taxa_df$Species == "Unassigned"]
+total_otus_greengenes <- sum(taxa_df_greengenes$Count)
+unassigned_idx_greengenes <- taxa_df_greengenes$Species == "Unassigned"
+no_species_assignment_greengenes <- taxa_df_greengenes$Count[unassigned_idx_greengenes]
+
+## Per genus count data - SILVA ################################################
+tidy_slv_taxa <- paeni_16S_silva %>% rowwise() %>%
+  mutate(new = str_split(Species, boundary("word")),Species = paste(new[1],new[2])) %>%
+  mutate(new = NULL) %>%
+  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
+  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
+  mutate(Species = str_remove(Species, "Paenibacillus "))
+
+taxa_df_silva <- tidy_slv_taxa %>%
+  group_by(Species) %>%
+  summarise(Count = n()) %>%
+  ungroup() %>%
+  mutate(Species = fct_reorder(Species, Count))
+
+## Count info for text
+total_otus_silva <- sum(taxa_df_silva$Count)
+unassigned_idx_silva <- taxa_df_silva$Species == "Unassigned"
+no_Species_assignment_silva <- taxa_df_silva$Count[unassigned_idx_silva]
+
+## Per genus count data - RDP ##################################################
+tidy_rdp_taxa <- paeni_16S_rdp %>% rowwise() %>%
+  dplyr::rename(Species = species, Genus = genus) %>%
+  mutate(new = str_split(Species, boundary("word")), Species = paste(new[1],new[2])) %>%
+  mutate(new = NULL) %>%
+  mutate(Species = if_else(!str_detect(Species, "Paenibacillus"), "Unassigned", Species)) %>%
+  mutate(Species = if_else(str_detect(Species, paste(c("Paenibacillus sp", "uncultured", "unidentified"), collapse = '|')), "Unassigned", Species)) %>%
+  mutate(Species = str_remove(Species, "Paenibacillus "))
+
+taxa_df_rdp <- tidy_rdp_taxa %>%
+  group_by(Species) %>%
+  summarise(Count = n()) %>%
+  ungroup() %>%
+  mutate(Species = fct_reorder(Species, Count))
+
+## Count info for text
+total_otus_rdp <- sum(taxa_df_rdp$Count)
+unassigned_idx_rdp <- taxa_df_rdp$Species == "Unassigned"
+no_Species_assignment_rdp <- taxa_df_rdp$Count[unassigned_idx_rdp]
+#print (paste0("Number of sequences with unassigned species label: ", taxa_df_rdp$Count[taxa_df_rdp$Species == "Unassigned"]))
 ```
 
-For the Greengenes 13.5 database there are a total of `r total_otus` sequences classified as `r nlevels(taxa_df$Species)` species in the Genus _Paenibacillus_. 
-The number of sequences assigned to specific _Paenibacillus_ species, range from 199 for _Paenibacillus amylolyticus_ to 2 for _Paenibacillus illinoisensis_ (Fig. \@ref(fig:speciesCount)). 
-Sequences only classified to the genus level, "Unassigned", is the most abundant group, `r no_species_assignment`. 
-```{r speciesCount, fig.cap = "Number of sequences assigned to species in the genus _Paenibacillus_."} 
-taxa_df %>% 
+- Relatively small overlap between the databases, multiple database analysis provide additional power.
+- The databases vary in the number of sequences for each species with some databases not having any sequences for some species.
+- Number of sequences and species for each database
+- Greengenes benefit - cleaner species names
+- SILVA benefit - additional sequences
+- RDP benefit - additional sequences
+- Issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_Species_assignment_silva`, and RDP - `r no_Species_assignment_rdp` __FIG-REF__.
+
+```{r speciesCount, echo = FALSE, fig.cap = "Number of sequences in the genus _Paenibacillus_ assigned to the 10 most represented species as well as sequences not assigned at the species level for the Greengenes, SILVA, and RDP databases.", message = FALSE}
+combined_taxa_df <- bind_rows(list(Greengenes = taxa_df_greengenes,
+                                   SILVA = taxa_df_silva,
+                                   RDP = taxa_df_rdp),
+                              .id = "Database") %>%
+  group_by(Species) %>%
+  mutate(max_count = max(Count, na.rm = TRUE)) %>%
+  ungroup() %>%
+  top_n(30, max_count) %>%
+  mutate(Species = fct_reorder(Species, max_count))
+
+combined_taxa_df %>%
+  group_by(Species) %>%
+  mutate(ymin = min(Count),
+         ymax = max(Count)) %>%
 ggplot() +
-    geom_bar(aes(x = Species, y = Count), stat = "identity") + 
-    geom_text(aes(x = Species, y = Count, label = Count),nudge_y = 75) +
-    labs(y = "Number of OTUs") + 
+    geom_linerange(aes(x = Species, ymin = ymin + 1, ymax = ymax + 1)) +
+    geom_point(aes(x = Species, y = Count + 1, color = Database)) +
+    labs(y = "Number of sequences") +
+  scale_y_log10() +
     coord_flip() +
     theme_bw() +
-    theme(axis.text.y = element_text(face = "italic"))
+    theme(axis.text.y = element_text(face = "italic"),
+          legend.position = "bottom")
 ```
 
-********************************************************************************
+## Taxonomic Resolution
+Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
+To differentiate between Species the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
+Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
 
-## Taxonomic resolution 
-Next we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ species by comparing within and between species amplicon pairwise distance for the V12 and V4 regions. 
-Inorder to differentiate between species the pairwise distances for within species amplicon seqences must be less than the between species distances. 
-Additionally the difference in amplicon sequence pairwise distances between and within species must be greater than the sequencing error rate to detect the difference. 
-For our taxonomic resolution analysis we use _in-silico_ PCR to extract the V12 and V4 regions of the 16S rRNA sequences. 
-We will then generate a pairwise distance matrix for the two regions and compare the within and between species pairwise distances.
+### Generating Distance Matrix for 16S Amplicons
+Here, we will demonstrate the process for extracting the target amplicon region and calculating pairwise distances using the Greengenes 13.5 database for simplicity.
+See the source Rmarkdown file for code used calculate pairwise distances for other databases and the V4 region as well as tidying the data for downstream analysis.
 
-For our _in-silico_ PCR we will use the following PCR primers:
+The V12 and V4 regions were extracted from the database sequences using pattern matching.
+Some sequences do not contain both forward and reverse primers as not all the  16S rRNA sequences are not full length.
+Only sequences with both forward and reverse primers are included in the analysis.
 
-|Region |Forward                    |Reverse                        |  
-|:------|:--------------------------|:------------------------------|  
-|V12    |27F - AGAGTTTGATCATGGCTCAG |336R - CACTGCTGCSYCCCGTAGGAGTCT|  
-|V4     |515F - GTGCCAGCMGCCGCGGTAA |806R - GGACTACHVGGGTWTCTAAT    |
+The following PCR primers were used for our _in-silico_ PCR:
 
-### V12 
-Extracting the V12 region from the database sequences, only sequences with the expected amplicon length of 318 basepairs are included in the analysis. 
-```{r}
+|Region |Direction |Primer                             |
+|:------|:-------- |:----------------------------------|
+|V12    |Forward   | 27F - AGAGTTTGATCATGGCTCAG        |
+|       |Reverse   | 336R - CACTGCTGCSYCCCGTAGGAGTCT   |
+|V4     |Forward   | 515F - GTGCCAGCMGCCGCGGTAA        |
+|       |Reverse   | 806R - GGACTACHVGGGTWTCTAAT       |
+
+```{r trimSeqs, echo = TRUE}
 forward_primer <- "AGAGTTTGATCATGGCTCAG"
 ## reverse complementing reverse primer
-reverse_primer <- as.character(reverseComplement(DNAString("CACTGCTGCSYCCCGTAGGAGTCT")))
+reverse_primer <- DNAString("CACTGCTGCSYCCCGTAGGAGTCT") %>%
+    reverseComplement() %>%
+    as.character()
 
-paeni_V12 <- TrimDNA(paeni_16S$seq, 
+## Selecting Paenibacillus taxa and seq
+paeni_16S_greengenes <- metagenomeFeatures::mgDb_select(gg13.5MgDb,
+                          type = c("taxa", "seq"),
+                          keys = "Paenibacillus",
+                          keytype = "Genus")
+
+## Finding sequeces with forward primer
+forward_match <- Biostrings::vmatchPattern(forward_primer,
+                                           subject = paeni_16S_greengenes$seq,
+                                           max.mismatch = 2) %>%
+    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")
+
+## Finding sequences with reverse primer
+reverse_match <- Biostrings::vmatchPattern(reverse_primer,
+                                           subject = paeni_16S_greengenes$seq,
+                                           max.mismatch = 2,
+                                           fixed = FALSE) %>%
+    as.list() %>% map_dfr(as.data.frame,.id = "seq_id")
+
+## sequences with both forward and reverse primers
+seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
+seqs_to_use <- names(paeni_16S_greengenes$seq) %in% seqs_to_use_ids
+
+## Trimming sequences with both primers
+paeni_V12 <- TrimDNA(paeni_16S_greengenes$seq[seqs_to_use],
                      leftPatterns = forward_primer,
@@
 
-seqs_to_use <- (width(paeni_V12[[2]]) == 318)
-v12_filtered <- paeni_V12[[2]][seqs_to_use]
+## Excluding seqs with length 0
+paeni_V12_seqs <- paeni_V12[[2]][width(paeni_V12[[2]]) != 0]
 ```
 
-Generating a multiple sequence alignment using the `AlignSeqs` function in the `DECIPHER` package.
-```{r}
+For more accurate distance estimates a multiple sequence alignment is used to calculate pairwise distances.
+We will use the `AlignSeqs` function in the `DECIPHER` package to generate the multiple sequence alignment.
 
-v12_align <- AlignSeqs(v12_filtered, verbose = FALSE)
+```{r align, echo = TRUE}
+v12_align <- AlignSeqs(paeni_V12[[2]], verbose = FALSE)
 ```
 
-The resulting alingment can be viewed using the `BrowseSeqs` function in the `DECIPHER` package.
-```{r eval = FALSE}
+The resulting alignment can be viewed using the `BrowseSeqs` function in the `DECIPHER` package.
+
+```{r eval = FALSE, echo = TRUE}
 BrowseSeqs(v12_align)
@@
 
-Generating pairwise distance matrix for taxonomic resolution analysis and converting distance matrix to data frame for analysis.
+Next a pairwise distance matrix is generated using the `DistanceMatrix` function in the `DECIPHER` package for taxonomic resolution analysis and converting distance matrix to a data frame for analysis.
 
-```{r}
+```{r v12Anno, echo = TRUE}
 v12_dist <- DistanceMatrix(v12_align,
@@
                            includeTerminalGaps = FALSE)
-
-v12_dist_df <- v12_dist %>% 
-    as.data.frame() %>% 
-    rownames_to_column(var = "Keys") %>% 
-    gather("Keys2","distance", -Keys) %>% 
-    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>% 
-    filter(Keys < Keys2) %>% 
-    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))
-
-tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species") 
-v12_dist_anno_df <- v12_dist_df %>% 
-    left_join(tax_df) %>% 
-    left_join(tax_df,by = c("Keys2" = "Keys")) %>% 
-    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>% 
-        mutate(group_comp = if_else(Keys_Species == Keys2_Species, "within","between")) %>% 
-    filter(Keys_Species != "s__", Keys2_Species != "s__")
 ```
 
-### V4 
-For the V4 region we will use the same approach, extract amplicon region, filter extracted sequences based on amplicon length, generate pairwise distance matrix using a multiple sequence alignment, and then evaluate pairwise distances.
+### Multiple Database Comparison
+Now that we have our pairwise distance matrix we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species.
+We will first compare the distribution of within and between species pairwise distances for the genus.
+Then look at the pairwise distances within species __XYZ__ compared distances between __XYZ__ and other _Paenibacillu_ species.
+To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the __###__ species with the most sequences.
+With more sequences we will have better estimates for within species pairwise distances but is likely to result in the exclusion of closely related species.
 
-```{r}
-## Extract amplicon region
-paeni_V4 <- TrimDNA(paeni_16S$seq, 
-                               leftPatterns = "GTGCCAGCMGCCGCGGTAA", 
-                               rightPattern = "ATTAGAWACCCBDGTAGTCC",
+```{r distDFfuns, results= "hide", echo = FALSE, message = FALSE}
+get_amp <- function(paeni_16S, forward_primer, reverse_primer){
+
+    ## Finding sequeces with forward primer
+    forward_match <- Biostrings::vmatchPattern(forward_primer,
+                                               subject = paeni_16S,
+                                               max.mismatch = 2) %>%
+        as.list() %>% map_dfr(as.data.frame,.id = "seq_id")
+
+    ## Finding sequences with reverse primer
+    reverse_match <- Biostrings::vmatchPattern(reverse_primer,
+                                               subject = paeni_16S,
+                                               max.mismatch = 2,
+                                               fixed = FALSE) %>%
+        as.list() %>% map_dfr(as.data.frame,.id = "seq_id")
+
+    ## sequences with both forward and reverse primers
+    seqs_to_use_ids <- intersect(forward_match$seq_id, reverse_match$seq_id)
+    seqs_to_use <- names(paeni_16S) %in% seqs_to_use_ids
+
+    ## Trimming sequences with both primers
+    paeni_amp <- TrimDNA(paeni_16S[seqs_to_use],
+                         leftPatterns = forward_primer,
+                         rightPatterns = reverse_primer,
                          type = "both")
 
-## Filter based on amplicon length
-v4_filtered <- paeni_V4[[2]][width(paeni_V4[[2]]) == 252]
+    ## Excluding seqs with lenght 0
+    paeni_amp <- paeni_amp[[2]][width(paeni_amp[[2]]) != 0]
 
-### Calculate distance matrix from multiple sequence alignment
-v4_align <- AlignSeqs(v4_filtered, verbose = FALSE)
-v4_dist <- DistanceMatrix(v4_align, 
+    ## Returning target seqs
+    paeni_amp
+}
+
+get_dist <- function(amp_seqs){
+  aligned_seqs <- AlignSeqs(amp_seqs, verbose = FALSE)
+  amp_dist <- DistanceMatrix(aligned_seqs,
                             correction = "none",
@@
 
+  Key_ord <- data.frame(Keys = rownames(amp_dist),
+                        Keys_ord = 1:nrow(amp_dist))
+  Key2_ord <- data.frame(Keys2 = rownames(amp_dist),
+                         Keys2_ord = 1:nrow(amp_dist))
+
   ## Creating a data frame for exploratory analysis
-v4_dist_df <- v4_dist %>% 
+  amp_dist_df <- amp_dist %>%
       as.data.frame() %>%
@@
       gather("Keys2","distance", -Keys) %>%
-    mutate(Keys = as.numeric(Keys), Keys2 = as.numeric(Keys2)) %>% 
-    filter(Keys < Keys2) %>% 
-    mutate(Keys = as.character(Keys), Keys2 = as.character(Keys2))
+      left_join(Key_ord) %>%
+      left_join(Key2_ord) %>%
+      filter(Keys_ord < Keys2_ord) %>%
+      select(-Keys_ord, -Keys2_ord)
+}
 
-tax_df <- dplyr::select(paeni_16S$taxa, "Keys", "Species") 
-v4_dist_anno_df <- v4_dist_df %>% 
+get_tidy_dist_df <- function(mgdb, tax_df, forward_primer, reverse_primer){
+    ## Get seqs and taxa
+    paeni_16S <- metagenomeFeatures::mgDb_select(mgdb,
+                              type = "seq",
+                              keys = tax_df$Keys,
+                              keytype = "Keys")
+
+    ## Subset seqs
+    amp_seqs <- get_amp(paeni_16S, forward_primer, reverse_primer)
+
+    ## Get distance DF
+    amp_dist_df <- get_dist(amp_seqs)
+
+    ## get distance df
+    amp_dist_anno_df <- amp_dist_df %>%
       left_join(tax_df) %>%
       left_join(tax_df,by = c("Keys2" = "Keys")) %>%
-    dplyr::rename(Keys_Species = Species.x, Keys2_Species = Species.y) %>% 
-    mutate(group_comp = if_else(Keys_Species == Keys2_Species, "within","between")) %>% 
-    filter(Keys_Species != "s__", Keys2_Species != "s__")
+      dplyr::rename(Keys_Species = Species.x,
+                    Keys2_Species = Species.y) %>%
+      mutate(group_comp = if_else(Keys_Species == Keys2_Species,
+                                  "within","between"))
 
+    ## Return tidy dataframe
+    amp_dist_anno_df
+}
+```
+
+```{r getDistDF, echo = FALSE, message = FALSE, results = "hide", warning = FALSE}
+### GG
+gg_for_seq_comp <- tidy_gg_taxa %>% 
+  filter(!is.na(rnacentral_ids), 
+         Species != "Unassigned", 
+         Species %in% combined_taxa_df$Species)
+
+### SILVA
+slv_for_seq_comp <- tidy_slv_taxa %>% 
+  filter(!is.na(RNAcentralID), 
+         Species != "Unassigned", 
+         Species %in% combined_taxa_df$Species)
+
+### RDP
+rdp_for_seq_comp <- tidy_rdp_taxa %>% 
+  filter(!is.na(RNAcentralID), 
+         Species != "Unassigned", 
+         Species %in% combined_taxa_df$Species)
+
+## V12 Amplicon
+forward_primer <- "AGAGTTTGATCATGGCTCAG"
+## reverse complementing reverse primer
+reverse_primer <- DNAString("CACTGCTGCSYCCCGTAGGAGTCT") %>%
+    reverseComplement() %>%
+    as.character()
+
+taxa_list <- list(Greengenes = gg_for_seq_comp,
+                  SILVA = slv_for_seq_comp,
+                  RDP = rdp_for_seq_comp)
+
+multidb_dist_V12_df <- map2_dfr(db_list, taxa_list,
+                                get_tidy_dist_df,
+                                forward_primer,
+                                reverse_primer,
+                                .id = "Database")
+
+## V4 Amplicon
+forward_primer <- "GTGCCAGCMGCCGCGGTAA"
+## reverse complementing reverse primer
+reverse_primer <- "ATTAGAWACCCBDGTAGTCC"
+
+multidb_dist_V4_df <- map2_dfr(db_list, taxa_list,
+                                get_tidy_dist_df,
+                                forward_primer,
+                                reverse_primer,
+                                .id = "Database")
+
+## Combine into a single dataframe
+multidb_dist_df <- bind_rows(
+  list(V12 = multidb_dist_V12_df,
+       V4  = multidb_dist_V4_df),
+  .id = "amplicon")
 ## Excluding outlier sequence "329842" - mean pairwise distance to all other
 ## sequences is 0.2
- v4_dist_anno_filt <- filter(v4_dist_anno_df,  Keys != "329842", Keys2 != "329842")
+multidb_dist_df <- filter(multidb_dist_df, Keys != "329842",Keys2 != "329842")
 ```
-
-********************************************************************************
-
-### Amplicon Sequence Lengths
-The trimmed sequence length ranges between 0 and 1500, due to sequences without primer matches (Fig. \@ref(fig:trimInfo)). 
-These are likey partial sequences or sequences with mutation in the primer binding regions. 
-The R package `PrimerMatch` (__REF__) can be used evaluate mismatches in the primer binding regions. 
-Here we will limit our taxonomic resolution analysis to trimmed sequences that are consistent with the expected amplicon length of 318 bp and 252 bp for the V12 and V4 regions repectively.
-
-```{r trimInfo, fig.cap = "Primer trimmed sequence length and start and end positions relative to full length sequences for the V12 and V4 regions.", fig.height = 4}
-list(V12 = paeni_V12[[1]], V4 = paeni_V4[[1]]) %>% 
-    map_dfr(as.data.frame, .id = "amplicon") %>% 
-    gather("key","value", -names, -amplicon) %>% 
-    mutate(key = if_else(key == "width", "length", key),
-           key = factor(key, levels = c("length","start","end"))) %>% 
-    mutate(position_type = if_else(key == "length","position", "length")) %>% 
-    ggplot() + 
-    geom_histogram(aes(x = value, fill = key), bins = 100) + 
-    facet_grid(position_type~amplicon, scales = "free") +
-    theme_bw()
-```
-
-********************************************************************************
 
@@
 
-Pairwise distance is significantly different for within and between species comparisons indicating that the V12 and V4 regions can potentially be used to classify members of the _Paenibacillus_ genus to the species level (Fig.
+Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable for classifying members of the _Paenibacillus_ genus to the Species level (Fig.
 \@ref(fig:pairDist)).
-Overall the V12 region had greater pairwise distances than V4 for both within and between species.
-It is important to also consider that the majority of sequences in the database were only classified to the genus level. 
-Species level information for these sequences might yield results that are inconsistent with our analysis.
+__TO ADD - Conclusions from multiple database analysis__.
 
-```{r pairDist, fig.cap = "Distribution of within and between species pairwise distances for the V4 16S rRNA region.
-Sequences not classified to the species level were excluded from the analysis.", fig.height = 3}
-list("V4" = v4_dist_anno_filt, "V12" = v12_dist_anno_df) %>% 
-    bind_rows(.id = "amplicon")  %>% 
-    ggplot(aes(x = group_comp, y = distance, fill = group_comp)) + 
-    geom_boxplot(aes(x = group_comp, y = distance, fill = group_comp)) + 
-      stat_compare_means(label.y = 0.2) +
-    facet_wrap(~amplicon) + 
-    labs(x = "Amplicon", fill = "Species Comparison") + 
+Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
+It is important also to consider that we only included sequences with species assignments and the species with the most sequences in the database.
+Species-level information for the unclassified sequences and inclusion of less well represented species might yield results that are inconsistent with our analysis.
+Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.
+
+```{r pairDist, fig.cap = "Distribution of within and between Species pairwise distances for the V12 and V4 16S rRNA region.
+Sequences not classified to the Species level were excluded from the analysis.", echo = FALSE}
+multidb_dist_df  %>%
+    ggplot(aes(x = Database, y = distance, fill = group_comp)) +
+    geom_boxplot() +
+    facet_grid(.~amplicon) +
+    labs(x = "Species Comparison", fill = "Species Comparison") +
         theme_bw() +
@@
 
-********************************************************************************
+#### Species level comparison
+Add text for new figure
 
-#### Species level comparison
-While the overall pairwise distance is greater between species than within species for the _Paenibacillus_ genus, it is important to understand how the within and between species pairwise distances compare for individual species.
-The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)). 
+<!--
+While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
+The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ Species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
 Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
-In general though for species levels classification the maximum within species distance should be less than the minimum between species distance. 
-For example as the maximum within species pairwise distance for _P.
-marcerans_ is 0.06 and the minimum between species pairwise distance for _P.
-marcerans_ and _P.
-barengoltzii_ is 0.02 (Fig.
-\@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two species is not possible.
+In general though for Species levels classification the maximum within Species distance should be less than the minimum between Species distance.
+For example as the maximum within Species pairwise distance for _P.
+lentimorbus_ is 0.13 and the minimum between Species pairwise distance for _P.
+lentimorbus_ and _P.
+alvei_ is 0.08 (Fig.
+\@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.
+-->
 
-```{r distHeatMap, fig.cap = "Pairwise distances between _Paenibacillus_ species (A) V12 and (B) V4 amplicon regions.
-Fill color indicates the mean pairwise sequence distance within and between species.
-The text indicates the maximum pairwise distance for within species comparisons, values along the diagonal, and maximum pairwise distance for between species comparisons.", fig.height = 8}
-v12_heat <- v12_dist_anno_df %>% 
-    group_by(Keys_Species, Keys2_Species, group_comp) %>% 
-    summarise(mean_dist = mean(distance),
-              max_dist = max(distance),
-              min_dist = min(distance)) %>% 
-    mutate(lab_dist = if_else(group_comp == "within", max_dist, min_dist)) %>% 
+```{r speciesComp, fig.cap = "Pairwise distances within the _Paenibacillus amylolyticus_  species and between  _P. amylolyticus_ and the other 9 most represented species.", echo = FALSE}
+multidb_dist_df %>%
+  mutate(Species_Comp = paste0(Keys_Species, Keys2_Species)) %>% 
+  filter(str_detect(Species_Comp, "amylolyticus")) %>% 
+  mutate(Species_Comp = str_remove(Species_Comp, "amylolyticus")) %>% 
+  mutate(Species_Comp = fct_reorder(Species_Comp, distance),
+         Species_Comp = fct_relevel(Species_Comp, "amylolyticus", after = 0)) %>% 
     ggplot() +
-    geom_raster(aes(x = Keys_Species, 
-                    y = Keys2_Species, 
-                    fill = mean_dist)) +
-    geom_text(aes(x = Keys_Species, y = Keys2_Species, label = round(lab_dist,2)), color = "white") + 
-    labs(x = "Species",y = "Species", fill = "Mean Dist") + 
+    geom_boxplot(aes(x = Species_Comp,
+                    y = distance,
+                    fill = Database)) +
+    facet_wrap(~amplicon, ncol = 1) +
    theme_bw() + 
-    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))
-
-v4_heat <- v4_dist_anno_df %>% 
-    filter(Keys != "329842", Keys2 != "329842") %>% 
-    group_by(Keys_Species, Keys2_Species, group_comp) %>% 
-    summarise(mean_dist = mean(distance),
-              max_dist = max(distance),
-              min_dist = min(distance)) %>% 
-    mutate(lab_dist = if_else(group_comp == "within", max_dist, min_dist)) %>% 
-    ggplot() + 
-    geom_raster(aes(x = Keys_Species, 
-                    y = Keys2_Species, 
-                    fill = mean_dist)) +
-    geom_text(aes(x = Keys_Species, y = Keys2_Species, label = round(lab_dist,2)), color = "white") + 
-    labs(x = "Species",y = "Species", fill = "Mean Dist") + 
-    theme_bw() + 
-    theme(axis.text.x = element_text(angle = -45, hjust = -0.1))
-
-ggpubr::ggarrange(v12_heat, v4_heat,labels = "AUTO",ncol = 1, nrow = 2,align = "v")
+  theme(axis.text.x = element_text(angle = -45, hjust = 0), 
+        legend.position = "bottom") + 
+  labs(x = "Species compared to amylolyticus", y = "Pairwise Distance")
 ```
 
-## Conclusion
-Here we demonstrate how the `metagenomeFeatures` package in conjunction with one of the associated 16S rRNA database packages, `greengenes13.5MgDb`, and other R packages, can be used to evaluate whether species level taxonomic classification is possible for a specific amplicon region.
-The approach used here can easily be extended to use different 16S rRNA databases (starting with a different `MgDb`clas object), taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
+# Conclusion
+__TO ADD - Statement about database overlap and using RNAcentralIDs__
+Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, _silva128.1MgDb_, _ribosomaldatabaseproject11.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
+The approach used here can easily be extended to use evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
 
@@
 
-## Session Information
-### System Information
-```{r}
+# Session Information
+## System Information
+```{r warning = FALSE}
 s_info <- devtools::session_info(include_base = FALSE)
@@
 ```
+
+<!-- __Potential reference__ https://www.biorxiv.org/content/early/2018/03/25/288654?%3Fcollection= -->
+<!-- Sequencing of the 16S ribosomal RNA (rRNA) gene and the fungal Internal Transcribed Spacer (ITS) region is widely used to survey microbial communities.
+Specialized ribosomal sequence databases have been developed to support this approach including Greengenes, SILVA and RDP.
+Most taxonomy annotations in these databases are predictions from sequence rather than authoritative assignments based on studies of type strains or isolates.
+Here, I investigate the error rates of taxonomy annotations in these databases.
+I found 253,485 sequences with conflicting annotations in SILVA v128 and Greengenes v13.5 at ranks up to phylum (9,644 conflicts), indicating that the annotation error rate in these databases is ~15%.
+I found that 34% of non-singleton genera have overlapping subtrees in the Greengenes tree from 2001 according to the RDP taxonomy, most of which are probably due to branching order errors in the Greengenes tree, which is therefore an unreliable guide to phylogeny.
+Using a blinded test, I estimated that the annotation error rate of the RDP database is ~10%.
+-->

### 2018-11-10 inst/manuscript/manuscript_supplemental.Rmd
text revisions

@@
 output: bookdown::pdf_document2
+bibliography: manuscript_supplemental.bib
 ---
@@
 With metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
-Here we demonstrate how _metagenomeFeatures_ and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
+Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
 Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s.
@@
 We used V12 and V4 region as they represent two commonly used amplicons for 16S rRNA marker-gene surveys.
-We will use the Greengenes 13.5 database, accessed using the _greengenes13.5MgDb_, Silva 128 database, accessed using the _silva128.1MgDb_, and RDP 11.5 database, accessed using the _ribosomaldatabaseproject11.5MgDb_, as annotation packages for our analysis of the _Paenibacillus_ genus.
+We will use the Greengenes 13.5 database, accessed using the `greengenes13.5MgDb`, Silva 128 database, accessed using the `silva128.1MgDb`, and RDP 11.5 database, accessed using the `ribosomaldatabaseproject11.5MgDb`, as annotation packages for our analysis of the _Paenibacillus_ genus.
 First, we show the distribution of sequences belonging to _Paenibacillus_ in these three database, and then a way to combine different subsets of database sequences to build a custom database object.
@@
 As a result the majority of the code used to produce this document is tidying the data.
-This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository, __ADD URL__
+This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository.[^1]
+
+[^1]: https://github.com/HCBravoLab/metagenomeFeatures/blob/master/inst/manuscript/manuscript_supplemental.Rmd
 
 # Required Packages
-In addition to _metagenomeFeatures_, _greengenes13.5MgDb_, _ribosomaldatabaseproject11.5MgDb_, and _silva128.1MgDb_, the _DECIPHER_, _tidyverse_, and _ggpubr_ packages are also used in the following analysis.
-Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@Wright2016-mo].
-The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse;@ggpubr]. 
-See __Session Information Section__ for package version information.
+In addition to `metagenomeFeatures`, `greengenes13.5MgDb`, `ribosomaldatabaseproject11.5MgDb`, and `silva128.1MgDb`, the `DECIPHER`, `tidyverse`, and `ggpubr` packages are also used in the following analysis.
+Our analysis uses the `DECIPHER` package to extract the amplicon regions, perform multiple sequence alignment, and generate a pairwise sequence distance matrix [@Wright2015-me].
+The `tidyverse` and `ggpubr` packages will be used to reformat the taxonomic and distance matrix data and generate summary figures [@tidyverse @ggpubr]. 
+See [Session Information Section](#sessionInfo) for package version information.
 
@@
 library(ggpubr)
-library(gplots)
+library(UpSetR)
 library(DECIPHER)
@@
 RNAcentral a meta-database for non-coding RNA sequences provides cross database unique identifiers for sequences in the databases. 
-The number of sequences in the databases and fraction of sequences with RNAcentralIDs varies for the three databases (__TABLE REF__). 
-Using the RNAcentral IDs we can evaluate the overlap between the three databases (__FIG VENN__).
+The number of sequences in the databases and fraction of sequences with RNAcentralIDs varies for the three databases (Table \@ref(tab:dbSize)). 
+Using the RNAcentral IDs we can evaluate the overlap between the three databases (Fig. \@ref(fig:dbVenn)).
 
-```{r echo = FALSE}
+```{r dbSize, echo = FALSE}
 get_rnacentral_entries <- function(mgdb){
@@
   dplyr::select(-n_entries_rnaid) %>% 
-  knitr::kable(digits = 2, format.args = list(big.mark = ","),
-               caption = "Number of sequences (size) and fraction of sequences with RNAcentral ids (frac ids) for the three 16S rRNA databases with MgDb annotation packages."
-               )
+  knitr::kable(digits = 2, format.args = list(big.mark = ","), booktabs = TRUE,
+               caption = "Number of sequences (size) and fraction of sequences with RNAcentral ids (frac ids) for the three 16S rRNA databases with MgDb annotation packages.")
 ```
@@
 ```{r dbVenn, echo = FALSE, fig.cap = "Overlap in sequence composition between the three 16S rRNA databases with  MgDb annotation packages for sequences with RNAcentral IDs."}
-venn(rna_central_id_list,)
+upset(fromList(rna_central_id_list), order.by = "freq")
 ```
@@
 no_Species_assignment_rdp <- taxa_df_rdp$Count[unassigned_idx_rdp]
-#print (paste0("Number of sequences with unassigned species label: ", taxa_df_rdp$Count[taxa_df_rdp$Species == "Unassigned"]))
 ```
 
-- Relatively small overlap between the databases, multiple database analysis provide additional power.
-- The databases vary in the number of sequences for each species with some databases not having any sequences for some species.
-- Number of sequences and species for each database
-- Greengenes benefit - cleaner species names
-- SILVA benefit - additional sequences
-- RDP benefit - additional sequences
-- Issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_Species_assignment_silva`, and RDP - `r no_Species_assignment_rdp` __FIG-REF__.
+For _Paenibacillus_ genus the databases vary in the number of sequences for each species with some databases not having any sequences for some species (Fig. \@ref(fig:speciesCount)). 
+With a relatively small overlap between the databases, multiple database analysis provide additional power  (Fig. \@ref(fig:dbVenn)). 
+The three database maintainers differ in their curation approaches, taxonomic name formating, and update frequency. 
+There are benefits to using all three databases in a cross database analysis. 
+The Greengenes database taxonomic name is more rigidly formated, only including the major taxonmic levels, e.g.
+does not include intermediate level such as suborder, and only goes to the species level.
+As a result species name parsing is unlikely to negatively impact the results. 
+SILVA and RDP are consistently updated, Greengenes has not been updated since 2013, therefore contain sequences that were not available in 2013, as a result the SILVA and RDP databases are significantly than Greengenes.
+Both the SILVA and RDP taxonomic name formating include intermediate taxonomic levels allowing for more fine grained taxonomic analysis. 
+Additionally, the SILVA database is the only one of the three to include eukaryotic 16S sequences as well as prokaryotic sequences.
+FOr our analysis of the  _Paenibacillus_ and issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_Species_assignment_silva`, and RDP - `r no_Species_assignment_rdp` (Fig.
+\@ref(fig:speciesCount)).
 
-```{r speciesCount, echo = FALSE, fig.cap = "Number of sequences in the genus _Paenibacillus_ assigned to the 10 most represented species as well as sequences not assigned at the species level for the Greengenes, SILVA, and RDP databases.", message = FALSE}
+<!--  Maybe table or upset plot for Paenibacillus - Number of sequences and species for each database   -->
+
+```{r speciesCount, echo = FALSE, fig.cap = "Number of sequences in the genus _Paenibacillus_ assigned to the 10 most represented species as well as sequences not assigned at the species level for the Greengenes, SILVA, and RDP databases.", message = FALSE, warning = FALSE}
 combined_taxa_df <- bind_rows(list(Greengenes = taxa_df_greengenes,
@@
 Now that we have our pairwise distance matrix we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species.
-We will first compare the distribution of within and between species pairwise distances for the genus.
-Then look at the pairwise distances within species __XYZ__ compared distances between __XYZ__ and other _Paenibacillu_ species.
-To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the __###__ species with the most sequences.
+To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the 10 species with the most sequences (Fig. \@ref(fig:speciesCount)).
 With more sequences we will have better estimates for within species pairwise distances but is likely to result in the exclusion of closely related species.
+We first compare the distribution of within and between species pairwise distances for the genus.
+Then look at the pairwise distances within species _P. amylolyticus_ compared distances between _P. amylolyticus_ and the other nine most represented _Paenibacillus_ species.
 
@@
 
-Pairwise distance is significantly different for within and between Species comparisons indicating that the V12 and V4 regions are potentially suitable for classifying members of the _Paenibacillus_ genus to the Species level (Fig.
+Pairwise distance is smaller within than between species, indicating that the V12 and V4 regions are potentially suitable for classifying members of the _Paenibacillus_ genus to the Species level (Fig.
 \@ref(fig:pairDist)).
-__TO ADD - Conclusions from multiple database analysis__.
+The median and interquartile pairwise distance range for within and between species comparisons in consistent across the three databases further supporting our observation. 
+A large number of outliers, pairwise distances past the boxplot wiskers, were observed for all three databases. 
+Outliers below the between between species boxplots and above the within species are problematic for species level classification. 
+The sequences and taxonomic assignment of these outliers should be evaluated to ensure they are not errors in the database, a result of incorrect database parsing, or are examples phylogeny and taxonomy are inconsistent.
 
@@
 
-```{r speciesComp, fig.cap = "Pairwise distances within the _Paenibacillus amylolyticus_  species and between  _P. amylolyticus_ and the other 9 most represented species.", echo = FALSE}
+```{r speciesComp, fig.cap = "Pairwise distances within the _Paenibacillus amylolyticus_  species and between  _P. amylolyticus_ and the 9 other most represented species.", echo = FALSE}
 multidb_dist_df %>%
@@
 # Conclusion
-__TO ADD - Statement about database overlap and using RNAcentralIDs__
-Here we demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, _greengenes13.5MgDb_, _silva128.1MgDb_, _ribosomaldatabaseproject11.5MgDb_, and other R packages, can be used to evaluate whether Species-level taxonomic classification is possible for a specific amplicon region.
+The `metagenomeFeatures` package and associated 16S rRNA database packages `greengenes13.5MgDb`, `silva128.1MgDb`,   `ribosomaldatabaseproject11.5MgDb`, provide a consistent interface for working with the different databases.
+Furthermore, through the use of the RNAcentralIDs we are able to evaluate database overlap and perform cross database analysis. 
+We demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, and other R packages, can be used to evaluate whether species-level taxonomic classification is possible for a specific amplicon region.
 The approach used here can easily be extended to use evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
@@
 
-# Session Information
+# Session Information {#sessionInfo}
 ## System Information
@@
 
+# References
 <!-- __Potential reference__ https://www.biorxiv.org/content/early/2018/03/25/288654?%3Fcollection= -->

### 2018-11-12 inst/manuscript/manuscript_supplemental.Rmd
Nidhi's edits

@@
 A limitation to 16S rRNA amplicon sequencing is a lack of taxonomic resolution, where organisms are only identifiable to the genus or family level.
-We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example differentiating between species within a genus.
+We define taxonomic resolution as the ability to differentiate between groups within a taxonomic level, for example, differentiating between species within a genus.
 We are interested in determining whether the 16S rRNA region of interest contains sufficient information for species-level taxonomic assignment.
@@
 
-There are multiple 16S rRNA databases whoes maintainers use different curation approaches resulting in different taxonomic and sequence compositions as well as being differently formatted. 
-Taxonomic or other 16S rRNA sequence analysis can perform analysis on an individual databases.
-With metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
+There are multiple 16S rRNA databases whose maintainers use different curation approaches resulting in different taxonomic and sequence compositions as well as being differently formatted. 
+One can perform taxonomic or other 16S rRNA sequence analysis on an individual databases but with metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
 Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
@@
 
-While the databases provide a consistent interface to the different 16S rRNA databases, the databases use different approaches to formating the taxonomic names.
-As a result the majority of the code used to produce this document is tidying the data.
-This code is not show in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository.[^1]
+While the databases provide a consistent interface to the different 16S rRNA databases, the databases use different approaches to formatting the taxonomic names.
+As a result, the majority of the code used to produce this document is tidying the data.
+This code is not shown in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository.[^1]
 
@@
 We have developed 16S rRNA MgDb annotation packages for three 16S rRNA databases, Greengenes, SILVA, and RDP. 
-RNAcentral a meta-database for non-coding RNA sequences provides cross database unique identifiers for sequences in the databases. 
+RNAcentral, a meta-database for non-coding RNA sequences provides cross database unique identifiers for sequences in the databases. 
 The number of sequences in the databases and fraction of sequences with RNAcentralIDs varies for the three databases (Table \@ref(tab:dbSize)). 
@@
 With a relatively small overlap between the databases, multiple database analysis provide additional power  (Fig. \@ref(fig:dbVenn)). 
-The three database maintainers differ in their curation approaches, taxonomic name formating, and update frequency. 
+The three database maintainers differ in their curation approaches, taxonomic name formatting, and update frequency. 
 There are benefits to using all three databases in a cross database analysis. 
-The Greengenes database taxonomic name is more rigidly formated, only including the major taxonmic levels, e.g.
+The Greengenes database taxonomic name is more rigidly formatted, only including the major taxonmic levels, e.g.
 does not include intermediate level such as suborder, and only goes to the species level.
-As a result species name parsing is unlikely to negatively impact the results. 
-SILVA and RDP are consistently updated, Greengenes has not been updated since 2013, therefore contain sequences that were not available in 2013, as a result the SILVA and RDP databases are significantly than Greengenes.
-Both the SILVA and RDP taxonomic name formating include intermediate taxonomic levels allowing for more fine grained taxonomic analysis. 
+As a result, species name parsing is unlikely to negatively impact the results. 
+SILVA and RDP are consistently updated while Greengenes has not been updated since 2013, and therefore it does not contain sequences that were not available in 2013.
+As a result, the SILVA and RDP databases are significantly more comprehensive than Greengenes.
+Both, the SILVA and RDP taxonomic name formating include intermediate taxonomic levels allowing for more fine grained taxonomic analysis. 
 Additionally, the SILVA database is the only one of the three to include eukaryotic 16S sequences as well as prokaryotic sequences.
-FOr our analysis of the  _Paenibacillus_ and issue with all databases, sequences only classified to the genus level, "Unassigned", is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_Species_assignment_silva`, and RDP - `r no_Species_assignment_rdp` (Fig.
+For our analysis of the  _Paenibacillus_, the and issue with all databases is that the sequences only classified to the genus level ("Unassigned" at species level) is the most abundant group, Greeengenes - `r no_species_assignment_greengenes`, SILVA - `r no_Species_assignment_silva`, and RDP - `r no_Species_assignment_rdp` (Fig.
 \@ref(fig:speciesCount)).
@@
 Next, we evaluate the 16S rRNA amplicon sequencing taxonomic resolution for _Paenibacillus_ Species by comparing within and between Species amplicon pairwise distance for the V12 and V4 regions.
-To differentiate between Species the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
-Additionally ,the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
+To differentiate between Species, the pairwise distances for within-Species amplicon sequences must be less than the between Species distances.
+Additionally, the difference in amplicon sequence pairwise distances between and within Species must be greater than the sequencing error rate to detect the difference.
 
@@
 Here, we will demonstrate the process for extracting the target amplicon region and calculating pairwise distances using the Greengenes 13.5 database for simplicity.
-See the source Rmarkdown file for code used calculate pairwise distances for other databases and the V4 region as well as tidying the data for downstream analysis.
+See the source Rmarkdown file for code used to calculate pairwise distances for other databases and the V4 region as well as tidying the data for downstream analysis.
 
 The V12 and V4 regions were extracted from the database sequences using pattern matching.
-Some sequences do not contain both forward and reverse primers as not all the  16S rRNA sequences are not full length.
+Some sequences do not contain both forward and reverse primers as not all the  16S rRNA sequences are full length.
 Only sequences with both forward and reverse primers are included in the analysis.
@@
 
-For more accurate distance estimates a multiple sequence alignment is used to calculate pairwise distances.
+For more accurate distance estimates, a multiple sequence alignment is used to calculate pairwise distances.
 We will use the `AlignSeqs` function in the `DECIPHER` package to generate the multiple sequence alignment.
@@
 ### Multiple Database Comparison
-Now that we have our pairwise distance matrix we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species.
-To reduce the complexity of our analysis we are only going to look at the pairwise distances between sequences assigned to the 10 species with the most sequences (Fig. \@ref(fig:speciesCount)).
+Now that we have our pairwise distance matrix, we can evaluate the taxonomic resolution of the genus _Paenibacillus_ by comparing the within and between species distances.
+To reduce the complexity of our analysis, we are only going to look at the pairwise distances between sequences assigned to the 10 species with the most sequences (Fig. \@ref(fig:speciesCount)).
 With more sequences we will have better estimates for within species pairwise distances but is likely to result in the exclusion of closely related species.
@@
 Outliers below the between between species boxplots and above the within species are problematic for species level classification. 
-The sequences and taxonomic assignment of these outliers should be evaluated to ensure they are not errors in the database, a result of incorrect database parsing, or are examples phylogeny and taxonomy are inconsistent.
+The sequences and taxonomic assignment of these outliers should be evaluated to ensure they are not errors in the database, a result of incorrect database parsing, or are examples where phylogeny and taxonomy are inconsistent.
 
-Overall the V12 region had greater pairwise distances than V4 for both within and between Species.
+Overall, the V12 region had greater pairwise distances than V4 for both within and between Species.
 It is important also to consider that we only included sequences with species assignments and the species with the most sequences in the database.
-Species-level information for the unclassified sequences and inclusion of less well represented species might yield results that are inconsistent with our analysis.
+Species-level information for the unclassified sequences and inclusion of less well-represented species might yield results that are inconsistent with our analysis.
 Additionally, our analysis does not identify the pairwise sequence distance required to classify a sequence as a novel _Paenibacillus_ Species.
@@
 We demonstrate how the _metagenomeFeatures_ package in conjunction with one of the associated 16S rRNA database packages, and other R packages, can be used to evaluate whether species-level taxonomic classification is possible for a specific amplicon region.
-The approach used here can easily be extended to use evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
+The approach used here can easily be extended to evaluate taxonomic groups (changing filtering parameters), or amplicon regions (changing primer sequences).
 

### 2018-11-13 R/mgDb_method_annotateFeatures.R
cleaning up namespace and documentation

@@ -5,12 +5,12 @@
 ################################################################################
 ## mgDb_annotateFeatures -------------------------------------------------------
 
-#' Generates an mgFeatures object when passed a vector of keys or dataframe
-#' (e.g. query)
-#' @param mgdb mdgdb object
-#' @param query search param
+# #' Generates an mgFeatures object when passed a vector of keys or dataframe
+# #' (e.g. query)
+# #' @param mgdb mdgdb object
+# #' @param query search param
 #' @importFrom dplyr right_join
-#' @keywords internals
+# #' @keywords internals
 .mgDb_annotateFeatures <- function(mgdb, query) {
     ##### SELECT_KEYS
     ## check query type: vector or data.frame
@@ -41,7 +41,7 @@
 
     ##### ANNOTATED_DB
     if (is.data.frame(query)) {
-        annotated_db <- right_join(query, filtered_db$taxa)
+        annotated_db <- dplyr::right_join(query, filtered_db$taxa)
     } else {
         annotated_db <- as.data.frame(filtered_db$taxa)
     }

### 2018-11-13 R/phylo-class.R

@@ -1,4 +1,4 @@
-## Setting phylo and phyloOrNULL class for MGDb and mgFeatures
+## Setting phylo and phyloOrNULL class for MgDb and mgFeatures
 
 setOldClass("phylo")
 

### 2018-11-13 R/mgDb-class.R
cleaning up namespace and documentation

@@ -7,10 +7,10 @@
 
 ## MgDb Class ------------------------------------------------------------------
 
-#' Tree can be either rds with phylo class object or tree file
-#' @param tree_file location of tree
+# #' Tree can be either rds with phylo class object or tree file
+# #' @param tree_file location of tree
 #' @importFrom ape read.tree
-#' @keywords internals
+# #' @keywords internals
 .load_tree <- function(tree_file){
     if (!file.exists(tree_file)) {
         stop(paste0("tree_file:", tree_file, " is not a valid file path"))

### 2018-11-13 R/mgDb-taxa_accessors.R

@@ -25,11 +25,12 @@ NULL
 
 
 ### Taxa keys function ---------------------------------------------------------
-#' Taxa keys functions
-#' import dplyr
-#' @param mgdb mgDb class
-#' @param keytype taxa keytype
-#' @keywords internals
+# #' Taxa keys functions
+#' @importFrom dplyr select_
+#' @importFrom dplyr collect
+# #' @param mgdb mgDb class
+# #' @param keytype taxa keytype
+# #' @keywords internals
 .taxa_keys <- function(mgdb, keytype){
     if (length(keytype) > 0) {
         nonvalid_keytype <- c()
@@ -68,9 +69,9 @@ setMethod("taxa_keys", "MgDb",
 
 
 ### Taxa columns function ------------------------------------------------------
-#' taxa columns
-#' @param mgdb mdgdb object
-#' @keywords internals
+# #' taxa columns
+# #' @param mgdb mdgdb object
+# #' @keywords internals
 .taxa_columns = function(mgdb){
     taxa_cols <- colnames(mgDb_taxa(mgdb))
     ## Excluding decipher column names
@@ -92,7 +93,7 @@ setMethod("taxa_columns", "MgDb",
 
 
 ### taxa keytypes function -----------------------------------------------------
-#' @keywords internals
+# #' @keywords internals
 .taxa_keytypes = function(mgdb){
     colnames(mgDb_taxa(mgdb))
 }

### 2018-11-13 R/mgDb_method_select.R

@@ -30,6 +30,7 @@
 # #' @param keys keys to filter
 # #' @param columns columns to select, defaults to all
 #' @importFrom dplyr select
+#' @importFrom dplyr select_
 #' @importFrom dplyr filter
 #' @importFrom dplyr collect
 #' @importFrom stringr str_c
@@ -96,7 +97,7 @@
 # #' @param keytype keytypes to select
 # #' @param keys keys to filter
 # #' @param columns columns to select, defaults to all
-# #' @import dplyr
+#' @importFrom dplyr select_
 # #' @keywords internals
 .select <- function(mgdb, type, keys, keytype, columns){
     ## check correct types
@@ -128,6 +129,9 @@
     # vector with all objects to return
     if ("taxa" %in% type || type == "all") {
         ## Removing decipher columns
+        ## avoiding namespace conflict with colname arguments in select
+        description <- NULL
+        identifier <- NULL
         select_obj$taxa <- dplyr::select(taxa_df, -row_names,
                                          -description, -identifier)
     }

### 2018-11-13 inst/manuscript/manuscript_supplemental.Rmd
edits

@@
 # Background
-16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance analysis.
+16S rRNA amplicon sequencing is commonly used for microbial community characterization, including differential abundance and diversity analysis.
 A limitation to 16S rRNA amplicon sequencing is a lack of taxonomic resolution, where organisms are only identifiable to the genus or family level.
@@
 
-There are multiple 16S rRNA databases whose maintainers use different curation approaches resulting in different taxonomic and sequence compositions as well as being differently formatted. 
-One can perform taxonomic or other 16S rRNA sequence analysis on an individual databases but with metagenomeFeatures and MgDb annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
-Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for a specific clade and amplicon region, specifically for the _Paenibacillus_ genus and V12 and V4 regions.
+There are multiple 16S rRNA databases each using their own file format and whose maintainers use different curation approaches resulting in different taxonomic and sequence compositions. 
+One can perform taxonomic or other 16S rRNA sequence analysis on an individual database but with `metagenomeFeatures` and `MgDb` annotation packages the same analysis can easily be performed on multiple databases, increasing the power of the analysis.
+Here we demonstrate how `metagenomeFeatures` and the `MgDb` annotation packages can be used to characterize taxonomic resolution for the _Paenibacillus_ genus and V12 and V4 amplicon regions.
 Originally, _Paenibacillus_ was classified under the _Bacillus_ genus, a novel genus was formed based on the 16S rRNA gene similarity in the 1990s.
@@
 We will use the Greengenes 13.5 database, accessed using the `greengenes13.5MgDb`, Silva 128 database, accessed using the `silva128.1MgDb`, and RDP 11.5 database, accessed using the `ribosomaldatabaseproject11.5MgDb`, as annotation packages for our analysis of the _Paenibacillus_ genus.
-First, we show the distribution of sequences belonging to _Paenibacillus_ in these three database, and then a way to combine different subsets of database sequences to build a custom database object.
+First, we show the distribution of sequences belonging to _Paenibacillus_ in these three databases, and then a way to combine different subsets of database sequences to build a custom database object.
 This helps investigators to pool sequences from different databases, to create an accurate and most representative database for their sample.
@@
 As a result, the majority of the code used to produce this document is tidying the data.
-This code is not shown in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures github repository.[^1]
+This code is not shown in the document but is available in the source Rmarkdown file which is provided as supplemental material and available in the metagenomeFeatures GitHub repository.[^1]
 
@@
 # 16S rRNA Database Comparison
-We have developed 16S rRNA MgDb annotation packages for three 16S rRNA databases, Greengenes, SILVA, and RDP. 
+We developed 16S rRNA MgDb annotation packages for three 16S rRNA databases, Greengenes, SILVA, and RDP. 
 RNAcentral, a meta-database for non-coding RNA sequences provides cross database unique identifiers for sequences in the databases. 
@@
 # _Paenibacillus_ Taxonomic Resolution
-For the taxonomic resolution analysis we will;
+For the taxonomic resolution analysis, we will:
 1. characterize the taxonomic composition of the _Paenibacillus_ genus in the three databases,
@@
 For _Paenibacillus_ genus the databases vary in the number of sequences for each species with some databases not having any sequences for some species (Fig. \@ref(fig:speciesCount)). 
-With a relatively small overlap between the databases, multiple database analysis provide additional power  (Fig. \@ref(fig:dbVenn)). 
+With a relatively small overlap between the databases, multiple database analysis provides additional power  (Fig. \@ref(fig:dbVenn)). 
 The three database maintainers differ in their curation approaches, taxonomic name formatting, and update frequency. 
 There are benefits to using all three databases in a cross database analysis. 
-The Greengenes database taxonomic name is more rigidly formatted, only including the major taxonmic levels, e.g.
+The Greengenes database taxonomic name is more rigidly formatted, only including the major taxonomic levels, e.g.
 does not include intermediate level such as suborder, and only goes to the species level.
@@
 As a result, the SILVA and RDP databases are significantly more comprehensive than Greengenes.
-Both, the SILVA and RDP taxonomic name formating include intermediate taxonomic levels allowing for more fine grained taxonomic analysis. 
+Both, the SILVA and RDP taxonomic name formatting include intermediate taxonomic levels allowing for more fine grained taxonomic analysis. 
 Additionally, the SILVA database is the only one of the three to include eukaryotic 16S sequences as well as prokaryotic sequences.
@@
 
-## Finding sequeces with forward primer
+## Finding sequences with forward primer
 forward_match <- Biostrings::vmatchPattern(forward_primer,
@@
 The median and interquartile pairwise distance range for within and between species comparisons in consistent across the three databases further supporting our observation. 
-A large number of outliers, pairwise distances past the boxplot wiskers, were observed for all three databases. 
-Outliers below the between between species boxplots and above the within species are problematic for species level classification. 
+A large number of outliers, pairwise distances past the boxplot whiskers, were observed for all three databases. 
+Outliers below the between species boxplots and above the within species are problematic for species level classification. 
 The sequences and taxonomic assignment of these outliers should be evaluated to ensure they are not errors in the database, a result of incorrect database parsing, or are examples where phylogeny and taxonomy are inconsistent.
@@
 #### Species level comparison
-Add text for new figure
-
-<!--
-While the overall pairwise distance is greater between Species than within Species for the _Paenibacillus_ genus, it is important to understand how the within and between Species pairwise distances compare for individual Species.
-The heatmap below shows pairwise distance information for within and between different _Paenibacillus_ Species for the V12 and V4 regions (Fig. \@ref(fig:distHeatMap)).
-Whether the sequences are assigned to more than one OTU depends on the pairwise sequence distance metric and linkage method employed by the clustering algorithm.
-In general though for Species levels classification the maximum within Species distance should be less than the minimum between Species distance.
-For example as the maximum within Species pairwise distance for _P.
-lentimorbus_ is 0.13 and the minimum between Species pairwise distance for _P.
-lentimorbus_ and _P.
-alvei_ is 0.08 (Fig.
-\@ref(fig:distHeatMap)A), correctly assigning a V12 amplicon sequences to one of these two Species is not possible.
--->
+While the overall pairwise distance is greater between than within species for the _Paenibacillus_ genus, it is important to understand how within and between species pairwise distances compare for individual Species.
+The within species pairwise for _P. amylolyticus_ distances are similar to between species distances for _P. pabuli_ and _P. illnoisensis_ (Fig. \@ref(fig:speciesComp)). 
+When including outliers the within pairwise distances for _P. amylolyticus_ is in the range of between pairwise distances for other _P. spp._.
+Similar to the genus level comparison investigation of the outliers sequences and strains is required to evaluate the taxonomic resolution for _Paenibacillus_.
 

### 2019-02-02 inst/manuscript/manuscript_supplemental.Rmd
fixed session info bug for submission

@@
 ## System Information
-```{r warning = FALSE}
-s_info <- devtools::session_info(include_base = FALSE)
-pander::pander(s_info$platform)
+```{r}
+sessioninfo::platform_info()
 ```
@@
 ```{r}
-s_info$packages %>% filter(`*` == "*") %>% select(-`*`) %>%
+sessioninfo::package_info() %>% 
+  filter(attached == TRUE) %>% 
+  select(package, loadedversion, source) %>% 
     knitr::kable(booktabs = TRUE)
